---
title: 'PLSC 40502: Problem Set 3'
author: "Solutions"
date: "February 20, 2022"
output:
  pdf_document: default
  html_document:
    df_print: paged
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Useful packages
library(tidyverse)
library(haven)
options(digits=3)
```


>This problem set is due at **11:59 pm on Friday, March 3rd**.

>Please upload your solutions as a .pdf file saved as "Yourlastname_Yourfirstinitial_pset3.pdf"). In addition, an electronic copy of your .Rmd file (saved as "Yourlastname_Yourfirstinitial_pset3.Rmd") must be submitted to the course website at the same time. We should be able to run your code without error messages. In addition to your solutions, please submit an annotated version of this `.rmd` file saved as "Yourlastname_Yourfirstinitial_pset3_feedback.rmd", noting the problems where you needed to consult the solutions and why along with any remaining questions or concerns about the material. In order to receive credit, homework submissions must be substantially started and all work must be shown.


# Problem 1

We'll continue with the same dataset from last problem set, the **Congressional Election Study**, an annual, large, nationally representative survey of the American population. We'll focus on the same outcome - predicting the level of support for the Section 232 Steel and Aluminum tariffs. 

Be very careful in reading the variable names and definitions (feel free to use your code from last problem if it differed from the code in this problem).

The code below will load in the Common Content from the 2020 CES. Please download the file directly from `https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi%3A10.7910/DVN/E9N6PH` and place it into your data folder:

```{r, warning = F, message = F}
ces <- read_csv("data/CES20_Common_OUTPUT_vv.csv")
```

The next few code fragments will carry out the relevant pre-processing to generate bins for all four of the covariates of interest.

```{r}

# State FIPS
ces$statefips <- str_pad(ces$inputstate, 2, pad="0")

# Outcome
ces <- ces %>% mutate(tariff = case_when(CC20_338b == "1" ~ 1,
                                         CC20_338b == "2" ~ 0,
                                         TRUE ~ NA_real_))

# Covariates
# Age
ces <- ces %>% mutate(age = 2020 - birthyr)
ces <- ces %>% mutate(age_bin = case_when(age>=18&age<=29 ~ "18-29",
                                          age>=30&age<=44 ~ "30-44",
                                          age>=45&age<=64 ~ "45-64",
                                          age>=65 ~ "65+"))

# Gender
ces <- ces %>% mutate(gender_bin = case_when(gender == 1 ~ "Male",
                                             gender == 2 ~ "Female"))

# Race
ces <- ces %>% mutate(race_bin = case_when(race==3|hispanic==1 ~ "Hispanic",
                                           race==1 ~ "White",
                                           race ==2 ~ "Black",
                                           race == 4  ~ "Asian",
                                           TRUE ~ "Other"))

# Education
ces <- ces %>% mutate(educ_bin = case_when(educ>=1&educ<=2 ~ "H.S. or less",
                                           educ>=3&educ<=4 ~ "Some college",
                                           educ==5 ~ "College degree",
                                           educ==6 ~ "Postgraduate"))
```

Make the variables factors with the correct baselines

```{r}

ces$gender_bin <- relevel(as.factor(ces$gender_bin), "Male")
ces$age_bin <- relevel(as.factor(ces$age_bin), "18-29")
ces$educ_bin <- relevel(as.factor(ces$educ_bin), "H.S. or less")
ces$race_bin <- relevel(as.factor(ces$race_bin), "White")

```



Subset down to the non-missing data

```{r}
ces_full <- ces %>% filter(!is.na(tariff)&!is.na(age_bin)&!is.na(race_bin)&!is.na(gender_bin)&!is.na(educ_bin)&!is.na(statefips))
```


## Part A

```{r}
library(brms)
library(rstan)

```


```{r, message=F, warning=F, results=F}
mlm_state_region <- brm(tariff ~ factor(gender_bin) + factor(age_bin) + factor(race_bin) + factor(educ_bin) +  (1 | region / statefips),
                           family = bernoulli(), data=ces_full, cores=4, warmup=1000, iter=2000)
```



```{r}
mlm_state <- brm(tariff ~ factor(gender_bin) + factor(age_bin) + factor(race_bin) + factor(educ_bin) +  (1 | statefips), 
                           family = bernoulli(), data=ces_full, cores=4, warmup=1000, iter=2000)
```



```{r}
ipums_frame <- read_csv("data/state_frame_2020_IPUMS.csv")
ipums_frame$gender_bin <- relevel(as.factor(ipums_frame$gender_bin), "Male")
ipums_frame$age_bin <- relevel(as.factor(ipums_frame$age_bin), "18-29")
ipums_frame$educ_bin <- relevel(as.factor(ipums_frame$educ_bin), "H.S. or less")
ipums_frame$race_bin <- relevel(as.factor(ipums_frame$race_bin), "White")
ipums_frame <- ipums_frame %>% left_join(read_csv("data/state_to_region.csv"), by="fips")

ipums_frame <- ipums_frame %>% rename(statefips=fips)
ipums_frame <- ipums_frame %>% group_by(statefips) %>% mutate(prop.state = count/sum(count)) %>% ungroup()
ipums_frame_predictions <- fitted(mlm_state_region, ipums_frame, summary=F)

ipums_frame$pm_model <- ipums_frame_prediction[,1]
```

```{r}

mrp_state <- ipums_frame %>% group_by(statefips) %>% summarize(tariffSupport = sum(pm_model*prop.state)*100) %>% ungroup() %>% rename(fips=statefips)

```

```{r}
library(usmap)
library(viridis)
```

```{r}
plot_usmap(regions="states", data= mrp_state, values="tariffSupport") + scale_fill_viridis(option = "A", name = "Tariff Support", label = scales::comma) +
  theme(legend.position = "right")
```

```{r}

us_2020 <- read_csv("data/usa_00002.csv")

#Recodes
us_2020 <- us_2020 %>% mutate(age_bin = case_when(AGE>=18&AGE<=29 ~ "18-29",
                                                  AGE>=30&AGE<=44 ~ "30-44",
                                                  AGE>=45&AGE<=64 ~ "45-64",
                                                  AGE>=65 ~ "65+"))

us_2020  <- us_2020 %>% mutate(gender_bin = case_when(SEX == 1 ~ "Male",
                                                      SEX == 2 ~ "Female"))

us_2020 <- us_2020 %>% mutate(race_bin = case_when((HISPAN==1|HISPAN==2|HISPAN==3|HISPAN==4) ~ "Hispanic",
                                                   RACE==1 ~ "White",
                                                   RACE ==2 ~ "Black",
                                                   RACE == 4| RACE == 5 | RACE== 6  ~ "Asian",
                                                   TRUE ~ "Other"))

us_2020 <- us_2020 %>% mutate(educ_bin = case_when(EDUCD == 999 ~ NA_character_,
                                                   EDUCD<=64 ~ "H.S. or less",
                                                   EDUCD>=65&EDUCD<=90~ "Some college",
                                                   EDUCD >= 100 & EDUCD <= 101 ~ "College degree",
                                                   EDUCD >= 110 ~ "Postgraduate"))

us_2020 <- us_2020 %>% select(STATEFIP, COUNTYFIP, PERWT, age_bin, gender_bin, race_bin, educ_bin)

state_data <- us_2020  %>% group_by(STATEFIP, age_bin, gender_bin, race_bin, educ_bin) %>% summarize(count = sum(PERWT))

state_data$fips <- str_pad(state_data$STATEFIP, 2, pad="0")

state_data %>% ungroup() %>% select(-STATEFIP) %>% write_csv("data/state_frame_2020_IPUMS.csv")

```
