---
title: "Week 1: Likelihood Inference"
subtitle: "PLSC 40502 - Statistical Models"
# author: "Anton Strezhnev"
output: 
  xaringan::moon_reader:
    self_contained: true
    css: [default, uchicago_pol_meth.css]
    nature:
      highlightLines: true
      
---
class: title-slide

# Welcome!
$$
  \require{cancel}
  \DeclareMathOperator*{\argmin}{arg\,min}
$$
  
```{r, echo = F, message=F, warnings=F}
library(tidyverse)
library(haven)
library(estimatr)
library(knitr)
options(digits=3)
```
---

# Course Overview

- Instructor: .maroon[Anton Strezhnev]
--

- Logistics:
  - Lectures T/Th - 2:00am - 12:20pm;
  - 3 Problem Sets (~ 2 weeks)
  - Final paper project (8-12 pages)
  - My office hours: Tuesdays 4pm-6pm (Pick 328)
--

- What is this course about?
  - *Defining* statistical models via their data-generating process
  - *Estimating* model parameters and conducting *inference*
  - *Interpreting* model output and *evaluating* model quality
--

- Goals for the course
  - Give you the tools you need to understand descriptive inference via statistical models and comment 
  on other researchers' work.
  - Equip you with an understanding of the fundamentals of likelihood and Bayesian inference
  to enable you to learn new models that build on these principles.
  - Teach you how to program and implement estimators by yourself!

---

$$
  \DeclareMathOperator*{\argmax}{arg\,max}
$$

# Course workflow 

- .maroon[Lectures]
  - Topics organized by week
  - Tuesday lecture: Introduction + higher-level overview
  - Thursday lecture: More in-depth details + applications
    - Lectures are the ``course notes" -- readings are the reference manuals.
--
- .maroon[Readings]
  - Mix of textbooks and papers
  - All readings available digitally on .maroon[Canvas]
  
---

# Course workflow

- .maroon[Problem sets] (35% of your grade)
  - Meant as a check on your understanding of the material and a way of communicating
  with me about the course.
  - The goal is learning by doing. So I'm not grading on correctness but more 
  on effort.
  - Solutions **will be posted** a week after the problem set is released. You
  should work on the problem sets without them but check your work afterward.
  - In addition to your write-up, you will also submit a brief survey about which questions
  you found challenging and needed to consult the solutions for + any concepts that are still unclear
  - Collaboration is **strongly encouraged** -- you should ask and answer questions on our private class
  Stack Overflow (link + invitation in the syllabus).
  
---

# Course workflow

- .maroon[Final Project] (55% of your grade)
  - The main goal of this class is for you to develop an independent quantitative research project
  - The paper should be in the length and style of a research note (8-12 pages)
    - One well-motivated question + data and analysis (minimize the lit review!)
  - You can collaborate! (1-3 authors per paper).
  - See the syllabus for published examples of the style/method of a paper that fits the aims of this class.
    - Survey data is a great place to ask descriptive questions
    - But feel free to use other sources or ask different types of questions - just talk to me about it!
--
- .maroon[Final Project Timeline]
  - **February 2nd**: 1 page project memo due
  - **February 28th/March 2nd**: Research presentations in-class (10-15 min. talks + Q&A)
  - **March 9**: Final paper due

---

# Class Requirements

- **Overall**: An interest in learning and willingness to ask questions.
--

- Assume a background in intro probability and statistics 
  - You should be comfortable thinking about basic estimands/estimators + their properties
  - You should be able to interpret a confidence interval for (e.g.) a difference-in-means.
--

- You should also be familiar with linear regression
  - $\hat{\beta} = (\mathbf{X}^{\prime}\mathbf{X})^{-1}\mathbf{X}^{\prime}Y$ should be a familiar expression
  - You should know under what conditions it's unbiased for $E[Y|X]$, and under what conditions it's efficient.
--

- If you want some review, check out chapters 1-6 of "Regression and Other Stories" 

---

# A brief overview

- **Week 1-2:** Introduction to likelihood inference and GLMs
  - Concept of the likelihood, MLE as an estimator + asymptotic properties
  - Binary outcome models, count models, duration models
--

- **Week 3-4:** Bayesian Inference and Multilevel Models
  - Principles of Bayesian inference -- posteriors, priors, data
  - Quantities of interest: posterior means, credible intervals
  - Estimation via MCMC
  - Application to multilevel regression models
--

- **Week 5:** Survey data
  - Applying multilevel regression methods to survey data
  - Survey weighting to address non-random sampling.
--

- **Week 6:** Mixture Models
--

- **Week 7:** Item response theory and ideal point models
--

- **Week 8:** Penalized regression and model selection
--

- **Week 9:** Research presentations + miscellaneous
---
class: title-slide

# Defining a statistical model

---

# Regression review

- A very common goal in statistics is to learn about the conditional expectation function $\mathbb{E}[Y|X]$
  - $Y_i$: Outcome/response/dependent variable
  - $X_i$: Vector of regressor/independent variables 
- "How does the expected value of $Y$ differ across different values of $X$?"
--

- Suppose we observe $N$ paired observations of $\{Y_i, X_i\}$. 
  - How do we construct a "good" estimator of $\mathbb{E}[Y|X]$?
  - What assumptions do we have to make to get...consistency...unbiasedness...efficiency?
--

- Consider the ordinary least squares estimator $\hat{\beta}$ which solves the minimization problem:
  
$$\hat{\beta} = \argmin_b \ \sum_{i=1}^N (Y_i - X_ib)^2$$
--

- We can do some algebra and find a closed form solution for this optimization problem

$$\hat{\beta} = (\mathbf{X}^{\prime}\mathbf{X})^{-1}(\mathbf{X}^{\prime}Y)$$
---

# Regression review

- **Assumption 1**: Linearity 

 $$Y = \mathbf{X}\beta + \epsilon$$
--

- **Assumption 2**: Strict exogeneity of the errors

 $$E[\epsilon | \mathbf{X}] = 0$$
--

- These two imply:

  - Linear CEF
  
$$\mathbb{E}[Y|\mathbf{X}] = \mathbf{X}\beta = \beta_0 + \beta_1X_{1} + \beta_2X_{2} + \dotsc \beta_kX_{k}$$
--

- **Best case**: Our CEF is truly linear (by luck or we have a *saturated* model)
- **Usual case**: We're at least consistent for the *best linear approximation* to the CEF

---

# Regression review

- **Assumption 3**: No perfect collinearity
  - $\mathbf{X}^{\prime}\mathbf{X}$ is invertible
  - $\mathbf{X}$ has full column rank
--

- This assumption is needed for *identifiability* -- otherwise no unique solution to the least squares minimization problem exists!
--

- Fails when one column can be written as a linear combination of the others
  - Or when there are more regressors than observations $k > n$

---


# Regression review

- Under assumptions 1-3, our OLS estimator $\hat{\beta}$ is unbiased and consistent for $\beta$
- Let's do a quick proof for unbiasedness

$$\begin{align*}\hat{\beta} &= (\mathbf{X}^{\prime}\mathbf{X})^{-1}(\mathbf{X}^{\prime}Y)\\
 &= (\mathbf{X}^{\prime}\mathbf{X})^{-1}(\mathbf{X}^{\prime}(\mathbf{X}\beta + \epsilon))\\
 &= (\mathbf{X}^{\prime}\mathbf{X})^{-1}(\mathbf{X}^{\prime}\mathbf{X})\beta + (\mathbf{X}^{\prime}\mathbf{X})^{-1}(\mathbf{X}^{\prime}\epsilon)\\
 &= \beta + (\mathbf{X}^{\prime}\mathbf{X})^{-1}(\mathbf{X}^{\prime}\epsilon)
 \end{align*}$$

--

- Then we can obtain the conditional expectation of $\mathbb{E}[\hat{\beta} | \mathbf{X}]$

$$\begin{align*} \mathbb{E}[\hat{\beta} | \mathbf{X}] &= \mathbb{E}\bigg[\beta + (\mathbf{X}^{\prime}\mathbf{X})^{-1}(\mathbf{X}^{\prime}\epsilon) \bigg| \mathbf{X} \bigg]\\
&= \mathbb{E}[\beta | \mathbf{X}] + \mathbb{E}[(\mathbf{X}^{\prime}\mathbf{X})^{-1}(\mathbf{X}^{\prime}\epsilon) | \mathbf{X}]\\
&= \beta + (\mathbf{X}^{\prime}\mathbf{X})^{-1}\mathbf{X}^{\prime} \mathbb{E}[\epsilon | \mathbf{X}]\\
&= \beta + (\mathbf{X}^{\prime}\mathbf{X})^{-1}\mathbf{X}^{\prime}0\\
&= \beta
 \end{align*}$$

---

# Regression review

- Lastly, by law of total expectation 

$$\mathbb{E}[\hat{\beta}] = \mathbb{E}[\mathbb{E}[\hat{\beta}|\mathbf{X}]]$$

- Therefore 

$$\mathbb{E}[\hat{\beta}] = \mathbb{E}[\beta] = \beta$$
--

- Consistency requires us to show the convergence of $(\mathbf{X}^{\prime}\mathbf{X})^{-1}(\mathbf{X}^{\prime}\epsilon)$ to $0$ in probability as $N \to \infty$.
  - This actually requires *weaker* assumptions: $\mathbb{E}[\mathbf{X}^{\prime}\epsilon] = 0$ but not necessarily $\mathbb{E}[\epsilon | \mathbf{X}] = 0$.

--
- But what have we not assumed?
  - Anything about the distribution of the errors!

---

# Regression review

- **Assumption 4** - Spherical errors

$$Var(\epsilon | \mathbf{X}) = \begin{bmatrix}
\sigma^2 & 0 & \cdots & 0\\
0 & \sigma^2 & \cdots & 0\\
\vdots & \vdots & \ddots & \vdots \\
0 & 0 & \cdots & \sigma^2
\end{bmatrix} =  \sigma^2 \mathbf{I}$$

--

- Benefits
  - Simple, unbiased estimator for the variance of $\hat{\beta}$
  - Completes Gauss-Markov assumptions $\leadsto$ OLS is BLUE (Best Linear Unbiased Estimator)
--

- Drawbacks
  - Basically never is true

---

# Regression review

- Good news! We can relax homoskedasticity (but still keep no correlation) and do inference on the variance of $\hat{\beta}$

$$Var(\epsilon | \mathbf{X}) = \begin{bmatrix}
\sigma^2_1 & 0 & \cdots & 0\\
0 & \sigma^2_2 & \cdots & 0\\
\vdots & \vdots & \ddots & \vdots \\
0 & 0 & \cdots & \sigma^2_n
\end{bmatrix}$$

--

- "Robust" standard errors using the "sandwich" estimator - Consistent but not unbiased for the true sampling variance of $\hat{\beta}$
  - Extensions to allow for forms of correlation in the error terms (e.g. "clustering")

---

# Regression review

- **Assumption 5** - Normality of the errors

$$\epsilon | \mathbf{X} \sim \mathcal{N}(0, \sigma^2)$$
--

- Not necessary even for Gauss-Markov assumptions
- Not needed to do asymptotic inference on $\hat{\beta}$
  - Why? Central Limit Theorem!

--
- Benefits?
  - Finite-sample inference.
  
---

# Regression review

- What do we need for OLS to be consistent for the "best linear approximation" to the CEF?
  - Very little!
--

- What do we need for OLS to be consistent and unbiased for the conditional expectation function?
  - Truly linear CEF
  - But still no assumptions about the outcome distribution!
--

- What do we need to do inference on $\hat{\beta}$?
  - We almost never assume homoskedasticity because "robust" SE estimators are ubiquitous
  - Even some forms of error correlation are permitted ("cluster" robust SEs)
  - Sample sizes are usually large enough where Central Limit Theorem kicks in since $\hat{\beta}$ can be written as a weighted sum over $Y_i$. So normality isn't that important.

---

# Defining a statistical model

- In the regression setting we tried to make as few assumptions about the data-generating process as possible.
  - Our goal is just to estimate and conduct inference on $E[Y|X]$.
--

- But what if we wanted to make further probabilistic statements about other quantities beyond $\beta$?
  - (e.g.) Can we provide a distribution for $Y_{n+1}$, the "next" observation given $X_{n+1}$?
  - If we're willing to make more assumptions about the data-generating process, we can do a lot more!
--

- **Statistical models** specify the data-generating process in terms of *systematic* and *stochastic* components.
  - **Systematic** elements are functions known constants and unknown *parameters*
  - **Stochastic** elements are draws from probability distributions
--

- We will be primarily working with *parametric* models
  - The data will be assumed to come from a particular family of probability distributions
  - The "structure" of the model is assumed fixed (the number of parameters does not grow with the size of the data).

---

# The linear model

- It is common to see the linear model written in its fully parametric form.
- **Stochastic**:

$$Y_i \sim \text{Normal}(\mu_i, \sigma^2)$$

- **Systematic**:

$$\mu_i = X_i^{\prime}\beta$$
--

- What's assumed to be known?
  - $\mathbf{X}$
- What's assumed to have a particular distribution?
  - $Y$
- We are interested in estimating and conducting inference on the parameters: $\beta$ and (less importantly) $\sigma^2$.

---

# General model notation

- We can specify a broad set of models for $Y_i$ using this framework
- **Stochastic**

$$Y_i \sim f(\theta_i, \alpha)$$

- **Systematic**

$$\theta_i = g(X_i, \beta)$$

- What are these quantities?
  - $Y_i$ is a random variable
  - $f()$ denotes the distribution of that random variable
  - $\theta_i$ and $\alpha$ are parameters of that distribution
  - $g()$ is some function 
  - $X_i$ are observed, known constants (e.g. regressors)
  - $\beta$ are parameters of interest 
--

- We will spend some time with a particular class of models called "Generalized Linear Models" where the systematic component has the form

$$\theta_i = g(X_i^{\prime}\beta)$$

---

# Types of distributions

- **Normal**

```{r, echo=F, fig.width=5, fig.height = 3, fig.align="center"}
norm_curve <- ggplot() + xlim(-5, 5) + geom_function(fun=dnorm) + theme_bw() + ylab("Density") + xlab("Value")
norm_curve
```

--
- Continuous on an unbounded support $(-\infty, \infty)$
- Two parameters: Mean $\mu$ and Variance $\sigma^2$
--

- Probability Density Function (PDF)

$$f_N(x;\mu, \sigma) = \frac{1}{\sigma\sqrt{2\pi}} \exp\left\{- \frac{1}{2}\left(\frac{x - \mu}{\sigma}\right) \right\}$$

---

# Types of distributions

- **Poisson**

```{r, echo=F, fig.width=5, fig.height = 3, fig.align="center"}
pois_curve <- data.frame(x=0:10) %>% ggplot(aes(x=x)) + xlim(0, 10) + geom_point(aes(y=dpois(x, lambda=1))) + theme_bw() + ylab("Probability") + xlab("Value")
pois_curve
```

--
- Discrete, defined on the support of the natural numbers (positive integers + zero)
- Single parameter: Mean and Variance $\lambda$ 


- Probability Density Function (PDF)

$$f(x; \lambda) = \frac{\lambda^x\exp\{-\lambda\}}{x!}$$
---

# Types of distributions

- **Binomial**

```{r, echo=F, fig.width=5, fig.height = 3, fig.align="center"}
binom_curve <- data.frame(x=0:10) %>% ggplot(aes(x=x)) + xlim(0, 10) + geom_point(aes(y=dbinom(x, size=10, prob = .3))) + theme_bw() + ylab("Probability") + xlab("Value")
binom_curve
```

- Discrete, defined on the support of integers from $\{0, 1, 2, \dotsc, n\}$ (model the sum of repeated i.i.d. coin flips.)
- Two parameters: $p$ probability of success in $n$ trials
  - Special case where $n=1$ trials is typically called the "Bernoulli"
- Probability Density Function (PDF)

$$f(x; p, n) = {n \choose x} p^x(1-p)^{n-x}$$

---

class: title-slide

# Likelihood inference

---

# Learning about the unknown

- Suppose that I want to learn about an unobserved parameter from a sample of observations.
  - I *assume* a particular statistical model for the data
  - **Example**: I want to know the level of support for then-President Donald Trump in Wisconsin in 2020 using the 2020 CES
--

- **Frequentist** approach
  - Unobserved parameters are **fixed constants**
  - Data are **random variables**
--

- We want to construct an **estimator** that is a *function* of the data and which has desirable properties
  - **Unbiasedness**: The *expected value* of the estimator is equal to the target parameter
  - **Consistency**: As our sample size gets larger, the estimator converges (in probability) to the target parameter.
  - **Asymptotic normality**: In large samples, the distribution of our estimator is normal (ideally with a variance we can estimate as well!)
--

- Can we come up with a *generic* framework that yields a "good" estimator for a large class of statistical models?

---

# The Likelihood Function

- Consider data $\mathbf{y} = \{y_1, y_2, \dotsc, y_n\}$ which we assume comes from a known distribution with density $f(\mathbf{y} | \theta)$ with unknown parameter $\theta$
  - $f()$ could be Bernoulli, Normal, Poisson, Gamma, Beta, etc... -- in this setting it is a *known* distribution
  - $\theta$ is an *unknown* parameter that lies in some space $\Theta$ of possible values.
--

- The **likelihood function** is a function of $\theta$ that is evaluated at the observed values of $\mathbf{y}$

$$\mathcal{L}(\theta| \mathbf{y}) = f(\mathbf{y} | \theta)$$
--

- Given some *input* $\theta$, the likelihood function returns the density of the observed data evaluated at that value.

--

- The likelihood function is **not** a probability density
  - a pdf is a function that takes $x$ as an input
  - a likelihood is a function that takes $\theta$ as an input
  
---

# The Likelihood Function

- The likelihood function is **not** $f(\theta | \mathbf{y})$
  - That statement doesn't even make sense in a frequentist framework -- parameters are constant
--

- Even when we (later) move to a *Bayesian* framework $f(\theta | \mathbf{y}) \neq f(\mathbf{y} | \theta)$
--

- Remember Bayes' Rule:
  
$$f(\theta | \mathbf{y}) = \frac{f(\mathbf{y} | \theta)}{f(\mathbf{y})} \times f(\theta)$$
--

- Or alternatively, since $f(\mathbf{y})$ is just a normalizing constant, you'll see this written as:
  
$$\underbrace{f(\theta | \mathbf{y})}_{\text{posterior}} \propto \underbrace{f(\mathbf{y}|\theta)}_{\text{likelihood}} \times \underbrace{f(\theta)}_{\text{prior}}$$

---

# The Likelihood Function

- We often make the assumption that our data are **independently and identically distributed**
  - $y_i \underset{{\text{i.i.d}}}{\sim} f(y_i | \theta)$
--

- This allows us to factor the likelihood 

$$\mathcal{L}(\theta|\mathbf{y}) = f(\mathbf{y} | \theta) = \prod_{i=1}^n f(y_i | \theta)$$
--

- Since our eventual goal will be to find an *optimum* of the likelihood, we can apply any monotonic function to it since that preserves maxima/minima.
  - The main one we'll apply is the **logarithm**
  - Why? Because logs turn annoying-to-work-with products into easier-to-work-with sums!
--

- The log-likelihood is typically denoted $\ell(\theta| \mathbf{y})$

$$\ell(\theta|\mathbf{y}) = \log f(\mathbf{y}|\theta) = \sum_{i=1}^n \log f(y_i | \theta)$$
---

# MLE

- We want to come up with an estimator $\hat{\theta}$ for the parameter $\theta$
  - $\hat{\theta}$ is a function of the data (like a sample mean or OLS coefficient)
  - Is there a principled way to pick $\hat{\theta}$ that has provably "good" properties across a wide variety of models?
--

- The Maximum Likelihood Estimator (MLE) $\hat{\theta}$ is defined as the value of $\theta$ that yields the optimum value of the likelihood function

$$\hat{\theta} = \argmax_\theta \ \log f(\mathbf{y}|\theta)$$
--

- The MLE has some desirable properties:
  - Under some regularity conditions, the MLE $\hat{\theta}$ is consistent for the true parameter $\theta$
  - It's asymptotically normal: $\sqrt{n}(\hat{\theta}_n - \theta) \rightarrow \text{Normal}(0, \frac{1}{\mathcal{I}(\theta)})$
  - It's asymptotically efficient: It's variance approaches the Cramer-Rao lower-bound $\frac{1}{n\mathcal{I}(\theta)}$
  - It's invariant to reparametarization: If $\hat{\theta}$ is an MLE for $\theta$, then $g(\hat{\theta})$ is an MLE for $g(\theta)$
--
  
---

# Score

- Before illustrating consistency, we need to define some additional functions of the log-likelihood.
- The first is the **score** or the gradient of $\ell(\theta|\mathbf{y})$ with respect to $\theta$

$$\mathbf{S} = \frac{\partial}{\partial\theta} \log f(\mathbf{y}|\theta)$$

--

- With i.i.d. data, you'll often see the score written in terms of the score for an individual observation $i$

$$\mathbf{S}_i = \frac{\partial}{\partial\theta} \log f(y_i|\theta)$$
--

- At the value of the true parameter $\theta_0$, the expected value of the score is $0$ (under regularity conditions)

$$ \mathbb{E}\bigg[\frac{\partial}{\partial\theta} \log f(y_i|\theta_0)\bigg] = \frac{\partial}{\partial\theta} \mathbb{E}[\log f(y_i|\theta_0)] = 0$$

---

# Consistency

- With this definition of the score, we can show consistency of the MLE under i.i.d. observations.
- By the weak law of large numbers

$$\frac{1}{n} \sum_{i=1}^n \log f(y_i | \theta) \underset{p}{\rightarrow} \mathbb{E}[\log f(y_i | \theta)]$$
--

- The MLE is an optimizer of the left-hand side.
  - Therefore it converges in probability to the optimizer of the right-hand side
--

- And we just showed that the score is $0$ at the true value of $\theta$, denoted $\theta_0$
  - Therefore $\theta_0$ is an extremum of the right-hand side (under a few additional regularity conditions)
--

- Therefore the MLE $\hat{\theta}$ is consistent for the true value $\theta_0$
  
--

---

# Consistency

- There are two important conditions for consistency that can be violated in practice
- **Identifiability**
  - No "plateaus" in the log-likelihood.
  - Two different values of $\theta$ can't both maximize the log-likelihood
  
$$f(\mathbf{y}|\theta) \neq f(\mathbf{y} | \theta_0) \ \forall \ \theta \neq \theta_0$$ 
  
--

- **Fixed parameter space**
  - The dimensionality of $\theta$ stays fixed and does not depend on $n$
  - Violated (e.g.) in ideal point models (new legislators mean new ideal points)
---

# Information

- Having established consistency, we're interested in understanding the variance and distribution (asymptotically) of $\hat{\theta}$ 
- To do this, we need to define a quantity called the **information**: $\mathcal{I}_n$
- This is equivalent to the variance of the score at its optimum

$$\mathcal{I}_n = \mathbb{E}\left[\left(\frac{\partial}{\partial\theta} \log f(\mathbf{y}|\theta_0)\right)^2\right]$$
--

- We can show (under some more regularity conditions) that this is equivalent to the negative expectation of the second-order partial derivative (Hessian) of the log-likelihood

$$\mathcal{I}_n = -\mathbb{E}\left[\frac{\partial^2}{\partial\theta^2} \log f(\mathbf{y}|\theta_0)\right]$$
- **Intuitively**: Captures the curvature of the log-likelihood at its maximum

---

# Information

- You'll sometimes see the information written in terms of the information from a single observation

$$\mathcal{I} =  -\mathbb{E}\left[\frac{\partial^2}{\partial\theta^2} \log f(y_i|\theta_0)\right]$$

- Under our i.i.d. assumption, we can write

$$\mathcal{I_n} = n\mathcal{I}$$

--

- This matters a bit for how we write/show consistency of the MLE
  
---

# Information

- Under our i.i.d. assumption (+ regularity conditions), we can show not only consistency but convergence in distribution to a normal

$$\sqrt{n}(\hat{\theta} - \theta) \underset{d}{\rightarrow} \text{Normal}(0, \mathcal{I}^{-1})$$

--

- So in large samples, a good approximation of the variance of $\hat{\theta}$ is $\mathcal{I}_n^{-1}$
  - An additional result: The inverse of the information is the lowest possible variance of any unbiased estimator (**Cramér-Rao lower bound**) 
  - MLE achieves this bound asymptotically

--

- How do we do inference then? Same way as before -- plug in a consistent estimator of $\mathcal{I}_n^{-1}$.
  - Calculate the hessian of the log-likelihood at the MLE and take the inverse of its negative.

---

class: title-slide

# Example

---

# Example: Bernoulli

- Suppose I was interested in learning about the proportion of residents in Wisconsin who approved of then-President Donald Trump's performance in 2020.
  - I observe a sample of $n$ respondents and observe an approve $(y_i = 1)$ or disapprove $(y_i = 0)$ response.
  - Ignore, for this example, the sample weights and assume we have a true simple random sample from the target populations.
--

- The data generating process:
  - *Stochastic*: $y_i \sim \text{Bernoulli}(\pi)$
  - *Systematic*: $\pi \in (0, 1)$
--

- Let's derive the (log)likelihood!

---

# Example: Bernoulli

- Under our assumption of i.i.d. observations:

$$\mathcal{L}(\pi | \mathbf{y}) = f(\mathbf{y} | \pi) = \prod_{i=1}^n f(y_i | \pi)$$
--

- The log-likelihood is

$$\ell(\pi | \mathbf{y}) = \sum_{i=1}^n \log f(y_i | \pi)$$
--

- Since $y_i$ is Bernoulli, the PMF is:

$$f(y_i | \pi) = \pi^{y_i}(1-\pi)^{1-y_i}$$
--

- Plugging back into the log-likelihood

$$\ell(\pi | \mathbf{y}) = \sum_{i=1}^n \log \bigg(\pi^{y_i}(1-\pi)^{1-y_i}\bigg)$$

---

# Example: Bernoulli

- Log of the product is the sum of the logs

$$\ell(\pi | \mathbf{y}) = \sum_{i=1}^n \log \bigg(\pi^{y_i}\bigg) + \log\bigg((1-\pi)^{1-y_i}\bigg)$$
--

- Properties of logs: $\log(a^b) = b\log(a)$

$$\ell(\pi | \mathbf{y}) = \sum_{i=1}^n y_i \log (\pi) + (1-y_i)\log(1-\pi)$$
--

- We'll simplify this further later, but let's take a look at the shape of this likelihood function w.r.t. its input $\pi$

---

# Example: Bernoulli

- Read in the data

```{r, warning=F, message=F}
approval <- read_csv("data/cces2020_trump_approval_WI_TX.csv")
approval <- approval %>% filter(!is.na(trumpapprove)) # Drop missing
approval_WI <- approval %>% filter(inputstate == 55) # Subset down to Wisconsin
approval_TX <- approval %>% filter(inputstate == 48) # Also get TX for comparison
```
--

- Write the likelihood function in code

```{r}
bern_lik <- function(pi, y){
  return(sum(y*log(pi) + (1-y)*log(1-pi)))
}
```
--

- Quick summary of the data

```{r}
table(approval_WI$trumpapprove)
```

---

# Visualizing the likelihood

- For Wisconsin

```{r, echo=F, fig.width=8, fig.height=6, fig.align="center"}
data.frame(x=seq(0.01,.99,by=.01)) %>% mutate(loglik=map_dbl(x, bern_lik, y=approval_WI$trumpapprove)) %>% 
ggplot(aes(x=x,y=loglik)) + xlim(0,1) + geom_line(lwd=2) + theme_bw() + xlab(expression(pi)) + ylab("log-likelihood")
```

---

# Visualizing the likelihood

- For Wisconsin and Texas

```{r, echo=F, fig.width=7, fig.height=6, fig.align="center" }
data.frame(x=seq(0.01,.99,by=.01)) %>% mutate(loglik=map_dbl(x, bern_lik, y=approval_WI$trumpapprove)) %>% mutate(loglik2=map_dbl(x, bern_lik, y=approval_TX$trumpapprove)) %>%
ggplot(aes(x=x,y=loglik)) + xlim(0,1) + geom_line(lwd=2) + geom_line(aes(y=loglik2),lwd=2, col='red') + theme_bw() + xlab(expression(pi)) + ylab("log-likelihood") + annotate("text", x=.4, y=-4000, label = "TX", col="red") + annotate("text", x=.4, y=-1500, label = "WI")
```

---

# Numerical optimization

- Often the MLEs can be obtained via closed-form solutions, but in most applications, the answer has to be obtained numerically.
- R has a built-in numerical optimizer: `optim` that implements some standard algorithms
  - But *tons* of other packages: https://cran.r-project.org/web/views/Optimization.html
--

- Let's use `optim()` to calculate the MLE

```{r, warning = F}
# Pass our likelihood through to the optimizer
mle_wi_optim <- optim(.5, fn=bern_lik, 
                      y=approval_WI$trumpapprove,
                      method = "BFGS",
                      control=list(fnscale=-1),
                      hessian=T)
```

---

# Numerical optimization

- What's the output of `optim()`?

```{r}
mle_wi_optim
```

---

# Numerical optimization

- Our MLE point estimate is:

```{r}
point_mle <- mle_wi_optim$par
point_mle
```

--

- A consistent estimator of the asymptotic variance is the inverse of the negative hessian

```{r}
var_mle <- solve(-mle_wi_optim$hessian)
sqrt(var_mle) # Standard error
```

--

- So our asymptotic 95% CI is

```{r}
c(point_mle - abs(qnorm(.025))*sqrt(var_mle), 
  point_mle + abs(qnorm(.025))*sqrt(var_mle))
```

---

# Analytical optimization

- The Bernoulli case is pretty simple, so let's try to find an analytical expression for the MLE. We left off at

$$\ell(\pi | \mathbf{y}) = \sum_{i=1}^n y_i \log (\pi) + (1-y_i)\log(1-\pi)$$
--

- To find the optimum, we first take the partial derivative with respect to $\pi$ and set it equal to zero

$$\frac{\partial}{\partial\pi}\ell(\pi | \mathbf{y}) = \frac{\partial}{\partial\pi}\sum_{i=1}^n y_i \log (\pi) + (1-y_i)\log(1-\pi)$$
--

- Derivatives and sums

$$\frac{\partial}{\partial\pi}\ell(\pi | \mathbf{y}) = \sum_{i=1}^n \left[\frac{\partial}{\partial\pi} y_i \log (\pi)\right] + \left[\frac{\partial}{\partial\pi}(1-y_i)\log(1-\pi)\right]$$

---

# Analytical optimization

- Chain rule and product rules

$$\frac{\partial}{\partial\pi}\ell(\pi | \mathbf{y}) = \sum_{i=1}^n \frac{y_i}{\pi} - \frac{1-y_i}{1-\pi}$$
--

- Split into 3 terms and pull out constants

$$\frac{\partial}{\partial\pi}\ell(\pi | \mathbf{y}) = \frac{1}{\pi}\sum_{i=1}^n y_i + \frac{1}{1-\pi}\sum_{i=1}^n y_i - \frac{n}{1-\pi}$$
--

- Set equal to $0$ and solve for $\pi$

$$0 = \frac{1}{\hat{\pi}}\sum_{i=1}^n y_i + \frac{1}{1-\hat{\pi}}\sum_{i=1}^n y_i - \frac{n}{1-\hat{\pi}}$$

---

# Analytical optimization

- Rearrange terms

$$\frac{n}{1-\hat{\pi}} = \frac{1}{\hat{\pi}}\sum_{i=1}^n y_i + \frac{1}{1-\hat{\pi}}\sum_{i=1}^n y_i$$
--

- Multiply through by $1-\hat{\pi}$

$$n = \frac{1-\hat{\pi}}{\hat{\pi}}\sum_{i=1}^n y_i + \sum_{i=1}^n y_i$$
--

- Cancel terms

$$n = \frac{1}{\hat{\pi}}\sum_{i=1}^n y_i$$

---

# Analytical optimization

- Divide by $n$, multiply by $\hat{\pi}$

$$\hat{\pi} = \frac{1}{n} \sum_{i=1}^n y_i$$

--

- Check that the second derivative is negative
   - Here it's strictly negative over the domain of $\pi$

$$\frac{\partial^2}{\partial\pi^2}\ell(\pi | \mathbf{y}) = \sum_{i=1}^n - \frac{y_i}{\pi^2} - \frac{1-y_i}{(1-\pi)^2}$$
--

- All that work and it turns out the MLE is just the **sample mean**!
  - Won't always be the case, but here the MLE is also *unbiased* because of what we know about the sample mean.
  
---

# Conclusion

- **Statistical models**
  - Describe the data-generating process in terms of *systematic* and *stochastic* components
  - In a parametric model, we assume the data $\mathbf{y}$ come from a known *distribution* with unknown parameters $\theta$
--

- **Likelihood**
  - A function of $\theta$ evaluated at the observed data $\mathbf{y}$ equal to $f(\mathbf{y}|\theta)$
  - How "likely" are my observed results in a world where the true DGP parameter were $\theta$.
  - Not $f(\theta|\mathbf{y})$. That quantity only makes sense in the Bayesian context and requires us to formulate a prior $f(\theta)$
--

- **Maximum Likelihood Estimator**
  - A *technique* to come up with a **good** estimator in any case where we can write down a (well-behaved) likelihood function for a DGP
  - Still a "function" of the data, but that function has useful properties:
    - Consistency, Asymptotic Normality, Asymptotic Efficiency
    
---