---
title: "Week 9: Flexible Regression Pt. 2"
subtitle: "PLSC 40502 - Statistical Models"
# author: "Anton Strezhnev"
output: 
  xaringan::moon_reader:
    self_contained: true
    css: [default, uchicago_pol_meth.css]
    nature:
      highlightLines: true
      ratio: '16:9'
      
---
class: title-slide

# Review
$$
  \require{cancel}
  \DeclareMathOperator*{\argmin}{arg\,min}
$$
  
```{r, echo = F, message=F, warnings=F}
library(tidyverse)
library(haven)
library(estimatr)
library(knitr)
library(splines)
library(lubridate)

options(digits=3)

set.seed(60639)

```

---

# Previously

$$
  \DeclareMathOperator*{\argmax}{arg\,max}
$$

- **Flexible functional forms**
  - Model $E[Y|X] = f(x)$ as a linear combination of **basis functions**
  - Polynomial bases
  - Step-function bases
  - Splines as a compromise between the two!
- **Regularization**
  - Many regressors $\leadsto$ overfitting
  - Penalize fits with many *extreme* covariates -- pull estimates towards zero.
  - L2 penalty ("ridge regression") - equivalent to Bayesian normal regression
  - L1 penalty ("lasso") - sparse solutions w/ many coefficients set to zero.
---

# This week

- **Kernels**
  - "Kernel smoothing" - Kernels as a "localization" tool
  - "Kernel methods" - Kernels as regression inputs
  - Two *different* uses of the term/concept - but some interesting relations
- **Forests**
  - Tree-based models for classification and regression
  - Ensembles/averages of trees via **random forests**
  - Diagnostics via "variable smoothing"
  - Extensions to detecting causal heterogeneity
  
---
class: title-slide

# Kernel smoothing

---

# Example: Modeling Bike Rentals

- Recall again our **bikeshare usage** dataset
  - Fanaee-T, Hadi, and Gama, Joao, "Event labeling combining ensemble detectors and background knowledge", *Progress in Artificial Intelligence* (2013): pp. 1-15
--

```{r, warning=F, message=F}
bike <- read_csv("data/bikes_hour.csv")
bike_by_hour <- bike %>% group_by(hr) %>% summarize(cnt = mean(cnt))
```

---

# Example: Modeling Bike Rentals

- Our most intuitive approach to estimating the CEF was to just take the **average** for a given hour of the day

```{r, message = F, warning=F, fig.width=9, fig.height=6, fig.align="center", echo=F}
bike %>% ggplot(aes(x=hr, y=cnt)) + geom_point(alpha=.1) + geom_point(data=bike_by_hour, col="red", shape=15, size=5)  + xlab("Hour of the day") + ylab("Count of Users") +
  theme_bw() + theme(text = element_text(size = 16))     
```

---

# Kernel smoothing

- One downside - I can only evaluate the function at each discrete hour.
  - What would I predict is the usage at 7:45pm?
  - Intuitively, I would want to take the average of the "nearest data points"
--

- **Kernel smoothing** methods give us a straightforward, non-parametric approach to estimating $f(x) = E[Y|X=x]$ without making strong assumptions about the functional form of the CEF.
  - "weighted average" of "similar" observations
--

- We define the similarity between two points $x_0$ and $x$ by way of a **kernel function** $K_h()$

$$K_h(x_0, x) = D\bigg(\frac{|x - x_0|}{h(x_0)}\bigg)$$

- $D$ is our particular choice of kernel function and $h(x_0)$ is the "bandwidth" or "window size"
  - Often use a constant bandwidth that does not depend on the inputs $h(x_0) = h$
--

- The kernel function is **symmetric** and **non-negative** 


---

# Kernel smoothing

- Uniform kernel

$$D(u) = \begin{cases} \frac{1}{2} &\text{ if } |u| \le 1\\
0 & \text{ otherwise}\end{cases}$$
--

- Triangular kernel

$$D(u) = \begin{cases} 1 - |u| &\text{ if } |u| \le 1\\
0 & \text{ otherwise}\end{cases}$$
--

- Gaussian (or "radial basis") kernel

$$D(u) = \frac{1}{2\pi}\exp\bigg(-\frac{1}{2}u^2\bigg)$$

---

# Kernel smoothing

```{r}
uniform_kernel <- function(x) {
  ifelse(abs(x) <= 1, 1/2, 0)
}

triangular_kernel <- function(x) {
  ifelse(abs(x) <= 1, 1 - abs(x), 0)
}

gaussian_kernel <- function(x) {
  (1 / sqrt(2 * pi)) * exp(-0.5 * x^2)
}
```

---

# Kernel smoothing

```{r,  message = F, warning=F, fig.width=9, fig.height=6, fig.align="center", echo=F}
# Generate a sequence of x values
x_values <- seq(-3, 3, by = 0.01)

# Create a data frame with the kernel functions
kernels_df <- data.frame(
  x = rep(x_values, 3),
  y = c(sapply(x_values, uniform_kernel), 
        sapply(x_values, triangular_kernel), 
        sapply(x_values, gaussian_kernel)),
  kernel = rep(c('Uniform', 'Triangular', 'Gaussian'), each = length(x_values))
)

# Plot the kernel functions using ggplot2
ggplot(kernels_df, aes(x = x, y = y, color = kernel)) +
  geom_line(size = 1.2) +
  labs(title = "Kernel Functions",
       x = "u",
       y = "Density",
       color = "Kernel Type") +
  theme_bw() +
  theme(legend.position = "top")
```

---

# Nadaraya-Watson estimator

- For a set of $N$ observations indexed by $i$ $\{y_i, x_i\}$ and given a kernel function $K_h()$ and a chosen bandwidth $h$, the **Nadarya-Watson** estimator of $f(x) = E[Y|X = x]$ is:

$$\hat{f}(x) = \frac{\sum_{i=1}^n K_h(x - x_i)y_i}{\sum_{i=1}^n K_h(x - x_i)}$$

--

- Our best prediction for $Y$ at $x$ is the average of the observed $y_i$, weighted by the "closeness" of their corresponding $x_i$ to the input $x$.

--

- Some intuitions:
  - **Larger bandwidths** = more influence from "further" points (lower variance, higher bias)
  - **Smaller bandwidths** = only "close" points have meaningful weight (higher variance, lower bias)
  
---

# Example: Modeling Bike Rentals

- In R, local kernel smoothing methods are implemented in the package `np`

```{r}
library(np)
```

- We'll start with a uniform kernel and a bandwidth of 3

```{r}
kernel_reg <- npreg(cnt ~ hr, data=bike, ckertype="uniform", bws = 3, regtype='lc')
summary(kernel_reg)
```

---

# Example: Modeling Bike Rentals

- Generate predictions on a sequence of points and plot the line

```{r}
kernel_predict <- data.frame(hr = seq(0, 23, by=.01),
                             cnt = predict(kernel_reg, newdata=data.frame(hr = seq(0, 23, by=.01))))
```

---

# Example: Modeling Bike Rentals

```{r, message = F, warning=F, fig.width=9, fig.height=6, fig.align="center", echo=F}
bike %>% ggplot(aes(x=hr, y=cnt)) + geom_point(alpha=.1) + geom_point(data=bike_by_hour, col="red", shape=15, size=5)  +
  geom_line(data = kernel_predict, linewidth=2, col = "dodgerblue") + 
  xlab("Hour of the day") + ylab("Count of Users") +
  theme_bw() + theme(text = element_text(size = 16)) 
```

---

# Example: Modeling Bike Rentals

- What about if we use a gaussian kernel

```{r}
gaussian_kernel <- npreg(cnt ~ hr, data=bike, ckertype="gaussian", bws = 3, regtype='lc')
summary(gaussian_kernel)
```

```{r}
gaussian_predict <- data.frame(hr = seq(0, 23, by=.01),
                             cnt = predict(gaussian_kernel, newdata=data.frame(hr = seq(0, 23, by=.01))))
```

---

# Example: Modeling Bike Rentals

```{r, message = F, warning=F, fig.width=9, fig.height=6, fig.align="center", echo=F}
bike %>% ggplot(aes(x=hr, y=cnt)) + geom_point(alpha=.1) + geom_point(data=bike_by_hour, col="red", shape=15, size=5)  +
  geom_line(data = gaussian_predict, linewidth=2, col = "dodgerblue") + 
  xlab("Hour of the day") + ylab("Count of Users") +
  theme_bw() + theme(text = element_text(size = 16)) 
```

---

# Example: Modeling Bike Rentals

- What happens with a narrower bandwidth?

```{r}
gaussian_kernel_2 <- npreg(cnt ~ hr, data=bike, ckertype="gaussian", bws = 1, regtype='lc')
summary(gaussian_kernel_2)
```

```{r}
gaussian_predict_2 <- data.frame(hr = seq(0, 23, by=.01),
                             cnt = predict(gaussian_kernel_2, newdata=data.frame(hr = seq(0, 23, by=.01))))
```

---

# Example: Modeling Bike Rentals

```{r, message = F, warning=F, fig.width=9, fig.height=6, fig.align="center", echo=F}
bike %>% ggplot(aes(x=hr, y=cnt)) + geom_point(alpha=.1) + geom_point(data=bike_by_hour, col="red", shape=15, size=5)  +
  geom_line(data = gaussian_predict_2, linewidth=2, col = "dodgerblue") + 
  xlab("Hour of the day") + ylab("Count of Users") +
  theme_bw() + theme(text = element_text(size = 16)) 
```

---

# Local linear regression

- Instead of just making our prediction a weighted average, we can incorporate a least-squares adjustment
  - Fit a **line** rather than a **constant**
--

- Our prediction for $f(x)$ is defined as the solution to the **least-squares** problem in terms of intercept $\alpha(x)$ and slope $\beta(x)$

$$\sum_{i=1}^n K_h(x, x_i)[y_i - \alpha(x) - \beta(x) x_i]$$
--

- Let's fit this with `np` using a uniform kernel and a narrow bandwidth.

```{r}
uniform_local_linear <- npreg(cnt ~ hr, data=bike, ckertype="uniform", bws = 2, regtype='ll')
uniform_predict_ll <- data.frame(hr = seq(0, 23, by=.01),
                             cnt = predict(uniform_local_linear, newdata=data.frame(hr = seq(0, 23, by=.01))))
```

---

# Example: Modeling Bike Rentals

```{r, message = F, warning=F, fig.width=9, fig.height=6, fig.align="center", echo=F}
bike %>% ggplot(aes(x=hr, y=cnt)) + geom_point(alpha=.1) + geom_point(data=bike_by_hour, col="red", shape=15, size=5)  +
  geom_line(data = uniform_predict_ll, linewidth=2, col = "dodgerblue") + 
  xlab("Hour of the day") + ylab("Count of Users") +
  theme_bw() + theme(text = element_text(size = 16)) 
```

---

class: title-slide

# Kernel Methods

---

# Regression as optimization

- Throughout this course and previous courses, our goal has been to find some function $f(x)$ of inputs $x$ that acts as a "best predictor" of $Y$
  - We define some objective function $\mathcal{L}$ and attempt to minimize/maximize it
  - Least squares; maximum likelihood
--

- Given a set of $n$ inputs, $\{x_i, y_i\}$, we want to find a function that solves
  
$$\hat{f} = \argmin_{f} \mathcal{L}(\{x_i, y_i\}_{i=1}^n)$$
---

# Space of functions

- So far, we've placed severe constraints on the space of functions over which we search.

--

- **Linear functions**

$$f(x) = \beta_0 + \beta_1x_1 + \beta_2x_2 + \beta_3x_3 + \dotsc$$
--

- **Basis expansion**

$$f(X) = \sum_{m=1}^M \beta_m h_m(X)$$

--

- Can we be more flexible than this while still retaining computational feasibility?

---

# Space of functions defined by kernels

- "Kernel methods" consider searching over a space of functions defined implicitly by a positive-definite kernel function.
  - Example, the **Gaussian** kernel
  
$$K(x, x^\prime) = \exp\bigg(-\frac{||x - x^\prime||^2}{2\sigma^2}\bigg)$$
--

- We'll consider a search over a set of functions defined by a linear combination of kernels evaluated at the input points in our data

$$f(x) = \sum_{i=1}^N c_i K(x, x_i)$$

--

- What is this space of functions? How "flexible" is it?

---

# Reproducing Kernel Hilbert Spaces

- A positive semi-definite kernel function $K(x, x^\prime)$ corresponds to a space of functions known as a "reproducing kernel Hilbert space" (RKHS).
  - A choice of kernel implies a mapping $\phi(x)$ that transforms the input vector $x$ into a higher dimensional (possibly infinite dimensional) space.
--

  - This space is equipped with an inner product $\langle \cdot, \cdot\rangle$ such that for any two functions, the inner product is given by an evaluation of the kernel at the inputs
  
$$K(x, x^\prime) = \langle \phi(x), \phi(x^\prime)\rangle$$
--

- As a result, we can work in the space of $\phi$ without ever having to evaluate $\phi$ directly -- only its kernel.

---

# "The Kernel Trick"

- One important feature of optimization over functions defined over a reproducing kernel Hilbert space is that even if the functions in the space are **infinite dimensional** - solutions to optimization problems are necessarily **finite dimensional** and have the form of a linear combination of kernels.
- The **representer theorem** states that for minimization problems of the form:

$$\hat{f} = \argmin_{f} \mathcal{L}(\{x_i, y_i\}_{i=1}^n) + \lambda||f||^2$$

the solution is a finite sum of weighted kernel functions

$$\hat{f} = \sum_{i=1}^n \hat{c}_i K(x, x_i)$$

---

# Kernel Regularized Least Squares

- Intuitively, what the "kernel trick" allows us to do is design a good functional approximator by way of a conventional ridge regression estimator.
  - Relatively well-known in machine learning (particularly for classification) throughout the 2000s, but introduced to political science by **Hainmueller and Hazlett (2014)** and labeled **Kernel regularized least squares**
--

- Our regression matrix is constructed by creating a total of $N$ columns.
  - Each row of the "kernel regression" matrix is a linear combination of the kernel evaluated between $x_i$ and each observation in the data

$$f(x_i) = c_1K(x_i, x_1) + c_2K(x_i, x_2) + \dotsc c_n K(x_i, x_n)$$  
  
--

- The regression matrix is $N \times N$
  - We need to regularize! 

---

# Example: Interpolating spatial data

- A very common problem in spatial statistics is predicting an outcome at any given latitude/longitude coordinate given a finite set of observations at particular coordinates
  - e.g. What is the estimated average temperature given observations at a number of weather monitoring stations?
  - Finding the "best" interpolation between the observed points is known as "kriging" in the geostatistics literature
--

- We'll use a sample of U.S. weather stations as an example:

```{r}
library(sf)
library(geosphere)
```

---

# Example: Interpolating spatial data

- Start by reading in the U.S. shapefile

```{r}
us_shapefile <- read_rds("data/us_shapefile.rds") %>%
  filter(!(STUSPS %in% c("AK", "HI", "AS", "GU", "MP", "PR", "VI")))
```

- Next, read in the NOAA Climate Assessment Database monthly weather station data from January, 2025.

```{r}
weather <- read_csv("data/monthly_summary_202501_v2.csv") %>% filter(country == "UNITED_STATES")
weather <- weather %>% filter(atmp > -99999&ntmp > -99999) # Missing obs hard coded as -99999
weather <- weather %>% mutate(deviation = atmp - ntmp)
weather <- weather %>% filter(!(state %in% c("AK", "HI", "AS", "GU", "MP", "UM", "PR", "VI")))
```

---

# Example: Interpolating spatial data

- Convert to a plotable object and overlay on the map

```{r}
# Convert the stations to a plottable object
weather_sf <- st_as_sf(weather, coords = c("lon", "lat"), crs = st_crs(us_shapefile), remove=F)
```

---

# Example: Interpolating spatial data

```{r,  message = F, warning=F, fig.width=9, fig.height=4, fig.align="center", echo=F, dpi=150}
# Create the ggplot
ggplot(data = weather_sf) +
  # Plot the U.S. boundaries
  geom_sf(data = us_shapefile, fill = "white", color = "black") +
  # Plot the points
  geom_point(aes(x=lon, y=lat, colour = deviation)) +
  scale_colour_viridis_c() +  # You can change the color scale as needed
  labs(title = "CADB Stations",
       colour = "Deviation") +
  theme_minimal()
```

---

# Example: Interpolating spatial data

```{r,  message = F, warning=F, fig.width=9, fig.height=4, fig.align="center", echo=F, dpi=150}
# Create the ggplot
ggplot(data = weather_sf %>% filter(deviation < quantile(weather$deviation, .99))) +
  # Plot the U.S. boundaries
  geom_sf(data = us_shapefile, fill = "white", color = "black") +
  # Plot the points
  geom_point(aes(x=lon, y=lat, colour = deviation)) +
  scale_colour_viridis_c() +  # You can change the color scale as needed
  labs(title = "CADB Stations (dropped outliers)",
       colour = "Deviation") +
  theme_minimal()
```

---

# Example: Interpolating spatial data

- We'll make a grid of 50km squares on which to to evaluate our predictions

```{r, cache=T}
# Set boundaries for U.S. (approximate)
lat_min <- 24.396308  # Southernmost point (Florida)
lat_max <- 49.384358  # Northernmost point (Canada border)
lon_min <- -125.0     # Westernmost point (California)
lon_max <- -66.93457  # Easternmost point (Maine)

# Approximate 50 km in degrees (this will vary by latitude, but we'll use an average factor)
lat_interval <- 50 / 110.574  # 1 degree latitude ≈ 110.574 km
lon_interval <- 50 / (111.320 * cos(mean(c(lat_min, lat_max)) * pi / 180))  # Longitude interval varies with latitude

# Create a sequence of points for latitude and longitude
latitudes <- seq(lat_min, lat_max, by = lat_interval)
longitudes <- seq(lon_min, lon_max, by = lon_interval)

# Create a grid of all combinations of latitudes and longitudes
grid_points <- expand.grid(lat = latitudes, lon = longitudes)

# Convert the grid points to an 'sf' object
grid_points_sf <- st_as_sf(grid_points, coords = c("lon", "lat"), crs = st_crs(us_shapefile), remove=F)

# Filter grid points to keep only those within the U.S. boundary
grid_points_within_us <- st_intersection(grid_points_sf, us_shapefile)
```

# Example: Interpolating spatial data

- Simple linear regression seems to miss a lot of the geographic nuances.

```{r}
weather_linear <- lm(deviation ~ lat + lon + lat*lon, data=weather)
grid_points_within_us$linpredict <- predict(weather_linear, newdata=grid_points_within_us)
```

---

# Example: Interpolating spatial data

```{r,  message = F, warning=F, fig.width=9, fig.height=4, fig.align="center", echo=F, dpi=150}
# Create the ggplot
ggplot(data = grid_points_within_us) +
  # Plot the U.S. boundaries
  geom_sf(data = us_shapefile, fill = "white", color = "black") +
  # Plot the heatmap with grid squares
  geom_tile(aes(x = lon, y = lat, fill = linpredict), width = lon_interval, height = lat_interval) +
  scale_fill_viridis_c() +  # You can change the color scale as needed
  coord_sf(xlim = c(lon_min, lon_max), ylim = c(lat_min, lat_max), expand = FALSE) +
  labs(title = "Linear regression interpolation",
       fill = "Deviation") +
  theme_minimal()
```

---

# Kernel Regularized Least Squares

- What happens when we fit KRLS with latitude and longitude as the covariate inputs.

```{r}
library(KRLS)
```

- Estimate the model 
  - We use the default Hainmueller and Hazlett (2014) approach to setting the bandwidth equal to the number of dimensions in the data. 
  - Standard leave-one-out cross-validation of the penalty parameter

```{r krls, warning=F, cache=T}
krls_reg <- krls(X = model.matrix(~lat + lon, data=weather)[,-1],
                 y = weather$deviation,
                 whichkernel = "gaussian",
                 vcov=F,
                 derivative=F)
```

---

# Kernel Regularized Least Squares

- What do the inputs look like?

```{r}
dim(krls_reg$K)
head(krls_reg$K[,1])
head(krls_reg$K[,2])
head(krls_reg$K[,3])
```


---

# Kernel Regularized Least Squares

- What do the inputs look like?

```{r, fig.align="center"}
hist(krls_reg$K)
```


---

# Kernel Regularized Least Squares

- Predict using the KRLS regression for each of our grid squares.

```{r}
grid_points_within_us$krlspredict <- predict(krls_reg, newdata= model.matrix(~lat + lon, data=grid_points_within_us)[,-1])$fit
```

---

# Kernel Regularized Least Squares

```{r,  message = F, warning=F, fig.width=9, fig.height=4, fig.align="center", echo=F, dpi=150}
# Create the ggplot
ggplot(data = grid_points_within_us) +
  # Plot the U.S. boundaries
  geom_sf(data = us_shapefile, fill = "white", color = "black") +
  # Plot the heatmap with grid squares
  geom_tile(aes(x = lon, y = lat, fill = krlspredict), width = lon_interval, height = lat_interval) +
  scale_fill_viridis_c() +  # You can change the color scale as needed
  coord_sf(xlim = c(lon_min, lon_max), ylim = c(lat_min, lat_max), expand = FALSE) +
  labs(title = "Kernel Regularized Least Squares interpolation",
       fill = "Deviation") +
  theme_minimal()
```

---

# Kernel Regularized Least Squares

- What if we chose a smaller bandwidth for the Gaussian kernel?

```{r krls2, warning=F, cache=T}
krls_reg_2 <- krls(X = model.matrix(~lat + lon, data=weather)[,-1],
                 y = weather$deviation,
                 sigma = 1/5,
                 whichkernel = "gaussian",
                 vcov=F,
                 derivative=F)
```

---

# Kernel Regularized Least Squares

- Predict using the KRLS regression for each of our grid squares.

```{r}
grid_points_within_us$krlspredict2 <- predict(krls_reg_2, newdata= model.matrix(~lat + lon, data=grid_points_within_us)[,-1])$fit
```

---

# Kernel Regularized Least Squares

```{r,  message = F, warning=F, fig.width=9, fig.height=4, fig.align="center", echo=F, dpi=150}
# Create the ggplot
ggplot(data = grid_points_within_us) +
  # Plot the U.S. boundaries
  geom_sf(data = us_shapefile, fill = "white", color = "black") +
  # Plot the heatmap with grid squares
  geom_tile(aes(x = lon, y = lat, fill = krlspredict2), width = lon_interval, height = lat_interval) +
  scale_fill_viridis_c() +  # You can change the color scale as needed
  coord_sf(xlim = c(lon_min, lon_max), ylim = c(lat_min, lat_max), expand = FALSE) +
  labs(title = "Kernel Regularized Least Squares interpolation (smaller bandwidth)",
       fill = "Deviation") +
  theme_minimal()
```
