---
title: "Week 5: Surveys and Weighting"
subtitle: "PLSC 40502 - Statistical Models"
# author: "Anton Strezhnev"
output: 
  xaringan::moon_reader:
    self_contained: true
    css: [default, uchicago_pol_meth.css]
    nature:
      highlightLines: true
      ratio: '16:9'
      
---
class: title-slide

# Review
$$
  \require{cancel}
  \DeclareMathOperator*{\argmin}{arg\,min}
$$
  
```{r, echo = F, message=F, warnings=F}
library(tidyverse)
library(haven)
library(estimatr)
library(knitr)

options(digits=3)

```

---

# Previously

$$
  \DeclareMathOperator*{\argmax}{arg\,max}
$$

- **Stan**

--

- **Diagnosing model fit**

---

# This week

- **Survey weighting and post-stratification**
--

- **Combining multilevel regression and post-stratification**

---

---
class: title-slide

# Survey sampling

---

# Surveys

- A common task in survey research is estimating an unknown population parameter $\theta$ from a sample of respondents $i \in \{1, 2, 3, \dotsc, N\}$
  - What share of voters in Michigan plan to vote for Joe Biden in the 2020 election?
  - What is the share of adults who are vaccinated in each U.S. county?
  - What share of residents in Nevada plan to vote in the 2024 general election?
---

- Historically, two approaches to survey sampling design
  - **Quota sampling** - Define a set of known demographic targets and obtain samples to 
  - **Random sampling**
---

# Non-response

- Survey research


---

# Example: CCES 2020

- Let's dive in to the 2020 CCES. 

```{r, warning=F, message=F}
library(survey)
cces <- read_csv("data/CCES_subset.csv") %>% filter(!is.na(trumpApprove))
```

- We'll be using a subset of the outcome data, looking at the share of respondents who state that they strongly or somewhat approve of then-President Donald Trump.
--

- Start by making a `svydesign` object - we'll pretend that the sampling weights don't exist for now

```{r}
cces_surv_unwt <- svydesign(~1, weights = ~1, data=cces)
```

--

- In-sample, what's the proportion of respondents who approve of Trump?

```{r}
svymean(~trumpApprove, design=cces_surv_unwt)
```

---

# Example: CCES 2020

- How does this compare to the properly weighted mean?

```{r}
cces_surv_wt <- svydesign(~1, weights = ~commonweight, data=cces)
svymean(~trumpApprove, design=cces_surv_wt)
```

- Trump's approval is a few points higher after the weighting adjustment.

- From the CCES Guide:

> the completed cases were weighted to the sampling frame using entropy balancing. The 2019
ACS was used as the frame for weighting the common content and the team samples. The
CES sample was weighted to match the distributions of the 2019 ACS on gender, age, race,
Hispanic origin, and education level.
> The moment conditions included age, gender, education, race, plus their interactions. The
resultant weights were then post-stratified by age, gender, education, race, “born again"
status, voter registration status, 2016 Presidential vote choice, and 2020 Presidential vote
choice as needed.

---

# Population targets

- We'll be using the 2020 ACS 5-year Public Use Microdata Sample
  - Obtain the complete **joint** distribution of region, race, age, gender and education in the U.S.

```{r, message=F, warning=F}
acs_targets <- read_csv("data/ACS_2020_microdata_3cat.csv")
```


- What's the **marginal** distribution of education in the target population?

```{r}
acs_targets %>% group_by(educ_bin) %>% summarize(n = sum(Count)) %>% ungroup() %>% mutate(prop = n/sum(n))
```
---
# Population targets

- How does it compare to the **unweighted** marginal distribution in the sample?

```{r}
cces %>% group_by(educ_bin) %>% summarize(n = n()) %>% ungroup() %>% mutate(prop = n/sum(n))
```

---

# Population targets

- What about the **joint** distribution of gender and education?
--

- In the **population**

```{r, message=F}
acs_targets %>% group_by(educ_bin, gender_bin) %>% summarize(n = sum(Count)) %>% ungroup() %>% mutate(prop = n/sum(n))
```

---

# Population targets

- In the **sample**

```{r, message=F}
cces %>% group_by(educ_bin, gender_bin) %>% summarize(n = n()) %>% ungroup() %>% mutate(prop = n/sum(n))
```

---

# Post-stratification

- We can also construct the post-stratification weights manually
  - Start by calculating the population proportions in each bin

```{r}
acs_targets <- acs_targets %>% mutate(strata = str_c(gender_bin, educ_bin, age_bin, sep="-"),
                                      pop_proportion = Count/sum(Count))
```
--

- Do the same for the sample

```{r}
cces <- cces %>% mutate(strata = str_c(gender_bin, educ_bin, age_bin, sep="-"))
cces_strat <- cces %>% group_by(strata) %>% summarize(n=n()) %>% ungroup() %>% mutate(samp_proportion = n/sum(n))
```
--

- Join the datasets

```{r}
cces <- cces %>% left_join(acs_targets%>% select(strata, pop_proportion), by="strata")
cces <- cces %>% left_join(cces_strat %>% select(strata, samp_proportion), by="strata")
```

---

# Post-stratification

- Construct the post-stratification weights

```{r}
cces <- cces %>% mutate(postStratWt = pop_proportion/samp_proportion)
```

- Take the weighted average to estimate Trump Approval

```{r}
weighted.mean(cces$trumpApprove, cces$postStratWt)
```

---


# Population targets

- Did the weights equalize the distributions? Let's look again at the joint distribution of gender and education
--

- In the **population**

```{r, message=F}
acs_targets %>% group_by(educ_bin, gender_bin) %>% summarize(n = sum(Count)) %>% ungroup() %>% mutate(prop = n/sum(n))
```

--

- In the re-weighted **sample**

```{r, message=F}
cces %>% group_by(educ_bin, gender_bin) %>% summarize(n = sum(postStratWt)) %>% ungroup() %>% mutate(prop = n/sum(n))
```

---

# Post-stratification.

- Create the post-stratification weights using `postStratify` in `survey`

```{r}
cces_postStrat <- postStratify(cces_surv_unwt, strata=~gender_bin + age_bin + educ_bin,
                               population = acs_targets %>% select(gender_bin, age_bin, educ_bin, Freq = Count))
```

--

- Then use `svymean`

```{r}
svymean(~trumpApprove, cces_postStrat)
```

  
---

# Post-stratification

- Plotting our post-stratification weights against the actual CCES weights

```{r, warning=F, message=F, fig.align="center", fig.width=8, fig.height=5}
cces %>% ggplot(aes(x=postStratWt, y=commonweight)) + geom_point() + geom_smooth(method="lm") + theme_bw() +
  xlab("Post-stratification weight") + ylab("CCES 2020 Weight")
```

---

# Post-stratification

- Note that we can combine survey design weights with additional post-stratification weights.
  - For example, if we want to take a national survey and re-weight it to different demographic targets
--

- What happens if we combine our post-stratification weights with the CCES design weights

```{r}
cces_postStrat_wt <- postStratify(cces_surv_wt, strata=~gender_bin + age_bin + educ_bin,
                               population = acs_targets %>% select(gender_bin, age_bin, educ_bin, Freq = Count))
```

--

- Our estimated Trump approval is closer to that original 44 percent (as expected, since our post-stratification variables are a subset of all the covariates that go into the CCES weights)

```{r}
svymean(~trumpApprove, cces_postStrat_wt)
```

---

# Raking

- Sometimes we only know the marginals and not the joint distributions. We can still construct weights that get us balance on the *marginals*.
- Start by generating the marginal counts (in the real world, these would be **all** that we have).

```{r}
acs_marginals <- list()
acs_marginals[["gender_bin"]] <- acs_targets %>% group_by(gender_bin) %>% summarize(Freq = sum(Count))
acs_marginals[["age_bin"]] <- acs_targets %>% group_by(age_bin) %>% summarize(Freq = sum(Count))
acs_marginals[["educ_bin"]] <- acs_targets %>% group_by(educ_bin) %>% summarize(Freq = sum(Count))
```

--

- Make the raking design

```{r}
cces_rake_unwt <- rake(cces_surv_unwt, sample.margins = list(~gender_bin, ~age_bin, ~educ_bin), 
                       population.margins=acs_marginals)
cces$rakeWt <- weights(cces_rake_unwt) 
```

--

- In this case, we actually do about as well as when we have the full joint distribution!

```{r}
svymean(~trumpApprove, cces_rake_unwt)
```

---

# Raking

- Note that raking will **not** guarantee balance on the full joint distribution


- In the **population**

```{r, message=F}
acs_targets %>% group_by(educ_bin, gender_bin) %>% summarize(n = sum(Count)) %>% ungroup() %>% mutate(prop = n/sum(n))
```

---

# Raking

- In the re-weighted **sample**

```{r, message=F}
cces %>% group_by(educ_bin, gender_bin) %>% summarize(n = sum(rakeWt)) %>% ungroup() %>% mutate(prop = n/sum(n))
```

---