---
title: "Week 2: Generalized Linear Models"
subtitle: "PLSC 40502 - Statistical Models"
# author: "Anton Strezhnev"
output: 
  xaringan::moon_reader:
    self_contained: true
    css: [default, uchicago_pol_meth.css]
    nature:
      highlightLines: true
      ratio: '16:9'
      
---
class: title-slide

# Review
$$
  \require{cancel}
  \DeclareMathOperator*{\argmin}{arg\,min}
$$
  
```{r, echo = F, message=F, warnings=F}
library(tidyverse)
library(haven)
library(estimatr)
library(knitr)

options(digits=3)

```

---

# Previously

$$
  \DeclareMathOperator*{\argmax}{arg\,max}
$$

- **Likelihood**: $\mathcal{L}(\theta| \mathbf{Y}) \propto f(\mathbf{Y}| \theta)$
  - Function of $\theta$ (not itself a probability density)
  - Comparisons of likelihoods across different parameters capture notions of model "fit"
  
$$\lambda(\theta_1, \theta_2) = \frac{\mathcal{L}(\theta_1 | \mathbf{Y})}{\mathcal{L}(\theta_2 | \mathbf{Y})}$$

--

- **Frequentist** inference: Use the likelihood to find a "good" estimator for $\theta_0$: the MLE $\hat{\theta}$
  - $\hat{\theta}$ is consistent for $\theta_0$
  - It's asymptotically normal
  - It's (asymptotic) variance is the inverse Fisher Information
  
---

# This week

- **Generalized linear models**
  - What happens when $\mathbb{E}[Y_i|X_i] \neq X_i^{\prime}\beta$?
  - Can we retain the linear form but relate it to a function of the CEF?
--

- **Types of GLMs**
  - Binary outcome models (e.g. logistic)
  - Ordinal/Multinomial outcome models 
  - Count outcome models (e.g. Poisson)
  - Duration models (e.g. exponential)
--

- **Robust** inference
  - What happens when the GLM outcome distributions are wrong?
  - Can we still do valid inference for the CEF?

---
class: title-slide

# Intro to GLMs

---

# Motivation: Propensity Scores

- Researchers wanting to estimate causal effects from observational designs often use a *weighting* estimator to account for non-random treatment assignment.
  - Observe treatment $D_i$, confounders $X_i$
  - Need to estimate $Pr(D_i = 1 | X_i)$ to construct "inverse propensity of treatment weights"
--

- **Example**: .maroon[Keriakes et. al. (2000)] *"Abciximab provides cost-effective survival advantage in high-volume interventional practice"*
  - Abciximab, an anti-clotting drug, is often used during certain types of heart surgery to reduce bleeding risk.
  - Keriakes et. al. (2000) look at 1472 surgeries in Ohio Heart Health Center
  - Abciximab was administered *non-randomly* -- some types of patients more likely to receive the drug than others
--

- **Key problem** - With many continuous covariates, hard to estimate $Pr(D_i = 1 | X_i)$ non-parametrically
  - One solution: Assume a parametric *model* for $D_i$ 

---

# Generalized Linear Models

- Generalized linear models (GLMs) have three components:
  1. A parametric distribution on $Y_i|X_i$ ("stochastic component")
  2. A linear predictor: $\eta_i = X_i^{\prime}\beta = \beta_0 + \beta_1X_{i1} + \beta_2X_{i2} + \dotsc \beta_kX_{ik}$ ("systematic component")
  3. A link function $g()$ applied to the CEF $E[Y_i|X_i]$ that yields the linear predictor
  
  $$g(E[Y_i | X_i]) = \eta_i$$
--

  Alternatively, we can write the CEF in terms of the "inverse-link" $g^{-1}()$ applied to the linear predictor
  
  $$E[Y_i | X_i] = g^{-1}(\eta_i)$$
---

# Exponential Family

- The types of probability distributions permitted on $Y_i$ are quite general: the **exponential family**
  - This contains the **normal** as well as many other common distributions like the bernoulli, Poisson, exponential, etc...
--

- Exponential family distributions have density functions of the form

$$P(y|\theta) = h(y) \exp\bigg\{b(\theta) \cdot T(y) - A(\theta)\bigg\}$$
where $h(y)$, $A(\theta)$, $\eta(\theta)$ and $T(y)$ are known functions

--
- The key intuition: exponential distributions factorize in a convenient way
  - if $b(\theta) = \theta$, then the distribution is in "canonical" form
  - $T(y)$ is a "sufficient statistic"

---

# Example: Bernoulli

- Consider the bernoulli PMF

$$P(y_i|\pi_i) = \pi_i^{y_i}(1-\pi_i)^{1-y_i}$$

--

- Take the log, then the exponent

$$P(y_i|\pi_i) = \exp\bigg\{\log\bigg[\pi_i^{y_i}(1-\pi_i)^{1-y_i}\bigg]\bigg\}$$

--

- Properties of logs

$$P(y_i|\pi_i) = \exp\bigg\{y_i\log(\pi_i) + (1-y_i)\log(1-\pi_i)\bigg\}$$

---

# Example: Bernoulli

- Rearranging and using properties of logs again

$$P(y_i|\pi_i) = \exp\bigg\{y_i\log\bigg(\frac{\pi_i}{1-\pi_i}\bigg) + \log(1-\pi_i)\bigg\}$$

--

- So our exponential form is
  - $h(y) = 1$
  - $T(y) = y_i$
  - $A(\theta) = \log(1-\pi_i)$
  - $b(\theta) = \log\bigg(\frac{\pi_i}{1-\pi_i}\bigg)$
--

- Critically, this is where we get a good link function
  - The "canonical parameter" is $\log\bigg(\frac{\pi}{1-\pi}\bigg)$.
  - The "canonical link" is the function that equates this parameter with the linear predictor $X_i^{\prime}\beta = \log\bigg(\frac{\pi_i}{1-\pi_i}\bigg)$
  
---

# Logistic regression

- The "logit" or "logistic" GLM models the **log-odds** of a binary outcome as a function of the linear predictor $X_i^{\prime}\beta$

$$Y_i \underset{\text{i.i.d}}{\sim} \text{Bernoulli}(\pi_i)$$

$$E[Y_i | X_i] = Pr(Y_i = 1 | X_i) = \pi_i$$

$$\log\bigg(\frac{\pi}{1-\pi}\bigg) = X_i^{\prime}\beta$$
--

- Alternatively, this is written in terms of the "inverse-link" function (the logistic function) that relates $\pi_i$ to $g^{-1}(X_i^{\prime}\beta)$

$$\pi_i = \frac{\exp(X_i^{\prime}\beta)}{1 + \exp(X_i^{\prime}\beta)} = \frac{1}{1 + \exp(-X_i^{\prime}\beta)}$$

---

# Inverse-link functions

- The logistic function maps inputs on $\mathbb{R}$ to $(0, 1)$

```{r, echo=F, fig.width=5, fig.height = 3, fig.align="center"}
logistic <- function(x) 1/(1 + exp(-x))
logistic_curve <- ggplot() + xlim(-5, 5) + ylim(0,1) + geom_function(fun=logistic) + theme_bw() + ylab(expression(pi)) + xlab(expression(paste("X",beta))) + geom_vline(xintercept=0, lty=2) + ggtitle("Logistic function")
logistic_curve
```

- Another link is the "probit" whose inverse link is the Normal CDF

```{r, echo=F, fig.width=5, fig.height = 3, fig.align="center"}
normal_cdf <- ggplot() + xlim(-5, 5) + ylim(0,1) + geom_function(fun=pnorm) + theme_bw() + ylab(expression(pi)) + xlab(expression(paste("X",beta))) + geom_vline(xintercept=0, lty=2) + ggtitle("Normal CDF")
normal_cdf
```

---

# Inverse-link functions

- When do we use probit vs. logit?
  - Computational convenience!
  - Probit has some good properties for Bayesian inference
--

- Can you tell the difference?

```{r, echo=F, fig.width=6, fig.height = 4, fig.align="center"}
normal_logit <- ggplot() + xlim(-5, 5) + ylim(0,1) + geom_function(fun=pnorm, col="dodgerblue") + geom_function(fun=logistic, lty=2, col="indianred") + theme_bw() + ylab(expression(pi)) + xlab(expression(paste("X",beta))) + geom_vline(xintercept=0, lty=2) + ggtitle("Logistic vs. Normal CDF")
normal_logit
```


---

# Estimation

- We obtain an estimate of $\beta$, $\hat{\beta}$ using maximum likelihood.
- Our MLE estimator is:

$$\hat{\beta} = \argmax_\beta \ \log f(\mathbf{y} | \beta, \mathbf{X})$$
--

- Recall our score function is:

$$S(\beta) = \nabla \log f(\mathbf{y} | \beta, \mathbf{X}) = \begin{bmatrix} \frac{\partial}{\partial \beta_0} \log f(\mathbf{y} | \beta, \mathbf{X}) \\
\frac{\partial}{\partial \beta_1} \log f(\mathbf{y} | \beta, \mathbf{X}) \\
\vdots \\
\frac{\partial}{\partial \beta_k} \log f(\mathbf{y} | \beta, \mathbf{X})  \end{bmatrix}$$
--

- The likelihood is convex, so finding the maximum equates to solving for the value of $\beta$ that sets $S(\beta) = 0$
  - A common numerical method is **Newton-Raphson**

---

# Newton-Raphson

- An iterative algorithm starts at some initial guess $\hat{\beta}^{(0)}$ 
- Let $\hat{\beta}^{(t)}$ denote the "current" value of $\hat{\beta}$ and $\hat{\beta}^{(t+1)}$ our update -- we keep iterating until convergence.
--

- Our goal is to solve for a zero of $S(\beta)$
  - Let's do a first-order Taylor approximation around our current guess $S(\hat{\beta}^{(t)})$
  
$$S(\beta) \approx S(\hat{\beta}^{(t)}) + \nabla S(\hat{\beta}^{(t)}) \left(\beta - \hat{\beta}^{(t)}\right)$$
--

- What's $\nabla S(\hat{\beta}^{(t)})$? 
  - It's the Jacobian of the gradient...or the matrix of second-order partial derivatives of the log-likelihood...or the Hessian!
  - Denote it $\mathbf{H}(\hat{\beta}^{(t)})$

---

# Newton-Raphson

- Our next value of $\hat{\beta}$ is the value of $\beta$ that sets the score equal to zero 

$$0 =  S(\hat{\beta}^{(t)}) + \mathbf{H}(\hat{\beta}^{(t)}) \left(\hat{\beta}^{(t+1)} - \hat{\beta}^{(t)}\right)$$

--

- Multiply through by the inverse hessian

$$\hat{\beta}^{(t+1)} =  \hat{\beta}^{(t)} - \mathbf{H}^{-1}(\hat{\beta}^{(t)}) S(\hat{\beta}^{(t)})$$

--

- Recall that the negative inverse hessian is also the **Observed Fisher Information** 
  - An alternative algorithm, **Fisher Scoring** substitutes this for the expected Fisher Information
  - Often these update steps can be expressed as solutions to a **weighted least squares** optimization problem
--

- All of these and more are implemented in the `maxLik` R package, which we will be using this week.
  - Generally more "current" than `optim()` and includes some convenience functions.
  
---

# Example: Propensity Scores

- Let's estimate a logistic propensity score model for treatment in the .maroon[Keriakes et. al. (2000)] dataset

```{r, warning=F, message=F}
# Read in the dataset
pci <- read_csv("data/pci.csv")
```

--

- We want to predict treatment: `abcix` using a mix of discrete and continuous covariates...
  - `stent` - Coronary stent deployment; binary indicator
  - `height` - Height in centimeters; numeric integer
  - `female` - Female gender; binary indicator
  - `diabetic` - Diabetes mellitus diagnosis; binary indicator
  - `acutemi` - Acute myocardial infarction within the previous 7 days; binary indicator
  - `ejecfrac` - Left ejection fraction; numeric integer
  - `ves1proc` - Number of vessels involved in the patient’s initial PCI procedure; numeric integer
  
---

# Example: Propensity Scores

- Make the design matrix $\mathbf{X}$

```{r, warning=F, message=F}
X_mat <- model.matrix(abcix ~ stent + height + female + diabetic + acutemi + ejecfrac + ves1proc, data=pci)
head(X_mat) # View the top of the matrix
```
--

- Sample size

```{r}
n_obs <- nrow(X_mat)
n_obs
```

---

# Example: Propensity Scores

- From our logit link function, we have

$$\log\left(\frac{\pi}{1-\pi}\right) = X_i^{\prime}\beta $$

and 

$$1 - \pi_i = 1 -  \frac{1}{1 + \exp(-X_i^{\prime}\beta)} = \frac{\exp(-X_i^{\prime}\beta)}{1 + \exp(-X_i^{\prime}\beta)} = \frac{1}{1 + \exp(X_i^{\prime}\beta)} $$


- Let's write the log-likelihood:

$$\ell(\beta | \mathbf{y}, \mathbf{X}) = \sum_{i=1}^Ny_i X_i^{\prime}\beta + \log\bigg(\frac{1}{1 + \exp(X_i^{\prime}\beta)}\bigg)$$
$$\ell(\beta | \mathbf{y}, \mathbf{X}) = \sum_{i=1}^Ny_i X_i^{\prime}\beta - \log\bigg(1 + \exp(X_i^{\prime}\beta)\bigg)$$

---

# Example: Propensity Scores

- And let's put the log-likelihood into code (this returns a vector of the log-likelihood for each observation)

```{r, warning = F, message = F}
logit_loglik <- function(beta, Y, X){
  eta <- X%*%beta # linear predictor
  lik <- Y*eta - log(1+exp(eta))
  return(lik)
}
```

--

- Now let's optimize it to get the MLE

```{r, warning=F, message=F}
library(maxLik) # Maximum Likelihood library

logit_mle <- maxLik(logit_loglik, 
                      Y=pci$abcix,
                      X=X_mat,
                      start = rep(0, ncol(X_mat)),
                      method = "NR")
```

---

# Example: Propensity Scores

- What did we get?
```{r}
est <- coef(logit_mle) # Our optimization routine
names(est) <- colnames(X_mat)
est

# Compare to built-in R routine?
logit_Rglm <- glm(abcix ~ stent + height + female + diabetic + acutemi + ejecfrac + ves1proc, 
                  data=pci, family=binomial(link="logit"))
coef(logit_Rglm)

```

---

# Example: Propensity Scores

- Let's obtain our (asymptotic) variance-covariance matrix

```{r}
logit_vcov <- solve(-hessian(logit_mle))
```

- Square root of the diagonal is our SEs

```{r}
logit_SEs <- sqrt(diag(logit_vcov))
```


---

# Example: Propensity Scores

- Let's get our t-statistics and p-values

```{r}
results <- rbind(coef(logit_mle), logit_SEs, 
                     coef(logit_mle)/logit_SEs, 
                     2*pnorm(-abs(coef(logit_mle)/logit_SEs)))
colnames(results) <- colnames(X_mat)
rownames(results) <- c("Estimate", "Std. Error", "Test statistic", "p-value")
results
```

---

# Interpreting logit coefficients

- How do we interpret the $\beta$s substantively?

$$\log\bigg(\frac{\pi_i}{1-\pi_i}\bigg) = \beta_0 + \beta_1X_{i1} + \beta_2X_{i2} + \dotsc + \beta_k X_{ik}$$

--

- Take the partial derivative w.r.t. $X_{i1}$

$$\frac{\partial}{\partial X_{i1}} \log\bigg(\frac{\pi_i}{1-\pi_i}\bigg) =  \beta_1$$
--

- So $\beta_k$ captures the change in the log-odds for a one-unit change in $X_{k}$
  - Descriptively, it's the difference in log-odds between two observations that differ in $X_{k}$ by one unit.
---

# Interpreting logit coefficients

- On the "log-odds" scale, the change due to an increase in $X_{k}$ does not depend on the values of the other $X$ variables (unless we explicitly specify an interaction).
  - But thinking on the log-odds scale is hard! We think in terms of probabilities.
  - This additivity *does not hold* when we take $\frac{\partial}{\partial X_{i1}} \pi_i$ 
--

- Logit models *implicitly* encode interactions with respect to the CEF $\mathbf{E}[Y_i | X_i]$

---

# Interpreting logit coefficients

- **Remember**: A one-unit change in the linear predictor corresponds to *different* changes in probability depending on your baseline.

```{r, echo=F, fig.width=5, fig.height = 3, fig.align="center"}
logistic <- function(x) 1/(1 + exp(-x))
draw_points <- data.frame(x1 = c(0, 1), x2 = c(0, 1), y1 = c(0,0), y2 = logistic(c(0,1)))
draw_points2 <- data.frame(x1 = c(-5, -5), x2 = c(0, 1), y1 = logistic(c(0,1)), y2 = logistic(c(0,1)))
logistic_marginal <- ggplot() + xlim(-5, 5) + ylim(0,1) + geom_function(fun=logistic) + theme_bw() + ylab(expression(pi)) + 
  xlab(expression(paste("X",beta))) + geom_vline(xintercept=0, lty=2) + ggtitle("Marginal Effect at 0") +
  geom_segment(aes(x=x1, y=y1, xend=x2, yend=y2), data=draw_points, col="dodgerblue", lwd=2) +
  geom_segment(aes(x=x1, y=y1, xend=x2, yend=y2), data=draw_points2, col="indianred", lwd=2) 
  
logistic_marginal
```
--

```{r, echo=F, fig.width=5, fig.height = 3, fig.align="center"}
logistic <- function(x) 1/(1 + exp(-x))
draw_points_2 <- data.frame(x1 = c(2, 3), x2 = c(2, 3), y1 = c(0,0), y2 = logistic(c(2,3)))
draw_points_22 <- data.frame(x1 = c(-5, -5), x2 = c(2, 3), y1 = logistic(c(2,3)), y2 = logistic(c(2,3)))
logistic_marginal_2 <- ggplot() + xlim(-5, 5) + ylim(0,1) + geom_function(fun=logistic) + theme_bw() + ylab(expression(pi)) + 
  xlab(expression(paste("X",beta))) + geom_vline(xintercept=0, lty=2) + ggtitle("Marginal Effect at 2") +
  geom_segment(aes(x=x1, y=y1, xend=x2, yend=y2), data=draw_points_2, col="dodgerblue", lwd=2) +
  geom_segment(aes(x=x1, y=y1, xend=x2, yend=y2), data=draw_points_22, col="indianred", lwd=2) 
  
logistic_marginal_2
```

---

# Transformed quantities

- We have $\hat{\beta}$, but in the propensity score case, we really want $\hat{\pi_i} = Pr(D_i = 1 | X_i)$
  - Just apply the inverse link to get the quantity we want
  
$$\hat{\pi_i} = \frac{1}{1 + \text{exp}(-X_i^{\prime}\hat{\beta})}$$
--

- The function of the MLEs is the MLE of the function
  - So $\hat{\pi_i}$ is consistent for the true propensity scores (under our modeling assumptions)
--

- But what if we want to do inference on $\hat{\pi_i}$ or obtain a confidence interval?
  - How do we obtain $\text{Var}(\hat{\pi_i}) = \text{Var}(g(\hat{\beta}))$?

---

# Delta method

- We know $Var(\hat{\beta})$
  - Asymptotically, it's the inverse Fisher Information or the inverse negative Hessian of the log-likelihood.
--

- To get $Var(g(\hat{\beta}))$, let's start with a first-order Taylor approximation around the true value $\beta$

$$g(\hat{\beta}) \approx g(\beta) + [\nabla g(\beta)^{\prime}](\hat{\beta} - \beta)$$
--

- Take the variance 

$$Var(g(\hat{\beta})) \approx Var\bigg(g(\beta) + [\nabla g(\beta)^{\prime}](\hat{\beta} - \beta)\bigg)$$
--

- Expand the sum

$$Var(g(\hat{\beta})) \approx Var\bigg(g(\beta) + [\nabla g(\beta)^{\prime}]\hat{\beta} - [\nabla g(\beta)^{\prime}]\beta\bigg)$$

---

# Delta method

- Variance of a constant is zero

$$Var(g(\hat{\beta})) \approx Var\bigg([\nabla g(\beta)^{\prime}]\hat{\beta}\bigg)$$
--

- Pull out and "square" the constant. We now get an expression in terms of our original variance-covariance matrix

$$Var(g(\hat{\beta})) \approx [\nabla g(\beta)^{\prime}]Var(\hat{\beta})[\nabla g(\beta)]$$
--

- This approximation is actually exact asymptotically (the higher-order terms of the Taylor polynomial go to zero)
  - Use the usual plug-in estimator for the the gradient at the MLE
  
---

# Example: Propensity Score

- Let's determine the propensity of receiving treatment for a patient at the median covariate values 

```{r}
X_medians <- apply(X_mat, 2, median)
X_medians
```
--

- Construct our prediction function

```{r}
pred_prob <- function(beta, X){
  return(1/(1 + exp(-X%*%beta)))
}
pred_median <- pred_prob(coef(logit_mle), X_medians)
pred_median
```

---

# Example: Propensity Score

- We could solve for the gradient in terms of $\beta$ analytically, but there are plenty of convenient functions that will do this numerically

```{r}
pred_prob_gradient <- numericGradient(pred_prob, 
                                      t0 = coef(logit_mle),
                                      X = X_medians)
```

--

- Applying the delta method

```{r}
pred_prob_var <- pred_prob_gradient%*%logit_vcov%*%t(pred_prob_gradient) # R treats vectors as row-vectors not column vectors
```

--

- Making our 95% asymptotic CI for $\hat{\pi}(X)$

```{r}
c(pred_median - abs(qnorm(.025))*sqrt(pred_prob_var),
  pred_median + abs(qnorm(.025))*sqrt(pred_prob_var))
```

---

# "Monte Carlo" Delta Method

- Alternatively, we could approximate the (asymptotic) sampling distribution of $\hat{\pi}$ by:
  1. Sampling from the known asymptotic distribution of $\hat{\beta}$
  2. Passing each sampled $\beta$ through to our function $\pi = g(\beta)$ 
  3. Taking the variance of the simulated $\pi$s
--

- With many independent samples, this will get arbitrarily close to the true sampling variance of $\hat{\pi}$
  - King, Tomz and Wittenberg (2000) is essentially this idea
--

- Note that this is **not** bootstrapping.
  - We're using our existing estimator of $\widehat{Var(\hat{\beta})}$
  - Rather it's doing a "monte carlo" simulation instead of the delta method -- no need to take derivatives!

---

# "Monte Carlo" Delta Method

- Let's try it:

```{r}
set.seed(60637)
sim_betas <- MASS::mvrnorm(n=1e5, mu = coef(logit_mle), Sigma = logit_vcov)
sim_pi <- apply(sim_betas, 1, function(x) pred_prob(x, X=X_medians))
```

--

- How close are we?

```{r}
c(pred_median - abs(qnorm(.025))*sd(sim_pi),
  pred_median + abs(qnorm(.025))*sd(sim_pi))
```

---

class: title-slide

# "Latent variables" and logit

---

# Latent variables

- Another common way of formulating discrete outcome regressions is in terms of an unobserved continuous "latent" variable and an observation mechanism.
  - Instead of putting a distribution on $Y_i$, we put one on an unobserved $Y_i^*$
  - $Y_i$ is some function of $Y_i^*$ (usually an indicator function of some sort)
--

- For the logistic regression

$$Y_i = \mathbb{1}(Y_i^* \ge 0)$$
$$Y_i^* = X_i^{\prime}\beta + \epsilon_i$$
$$\epsilon_i \sim \text{Logistic}(0, 1)$$

--

- The Logistic distribution is parameterized in terms of a "location" (mean) $\mu$ and a "scale" parameter $s>0$. 
  - For $\mu = 0$, $s= 1$ we have the "standard" logistic

$$P(\epsilon_i < x) = \frac{1}{1 + \exp(-x)}$$

---

# Latent variables

- So what's $E[Y_i|X_i] = Pr(Y_i = 1|X_i)$?

$$Pr(Y_i = 1|X_i) = Pr(Y_i^* \ge 0) = Pr(\epsilon_i < X_i^{\prime}\beta) = \frac{1}{1 + \exp(-X_i^{\prime}\beta)}$$
--

- We can write the **probit** similarly, defining the error distribution as

$$\epsilon_i \sim \text{Normal}(0, 1)$$
--

- And by extension, the probability $Pr(Y_i|X_i)$

$$Pr(Y_i = 1|X_i) = Pr(Y_i^* \ge 0) = Pr(\epsilon_i < X_i^{\prime}\beta) = \Phi(X_i^{\prime}\beta)$$

---

# Latent variables

- We can visualize this as an integral over the density of the random "error" $\epsilon_i$

```{r, echo=F, fig.width=5, fig.height = 3, fig.align="center"}
logistic_dens <- ggplot() + xlim(-5, 5) + geom_function(fun=dlogis) + theme_bw() + ylab("Density") + geom_area(aes(x=c(-5,2)),stat = "function", fun = dlogis, fill = "dodgerblue", xlim = c(-5, 2)) + 
  xlab(expression(epsilon)) + geom_vline(xintercept=0, lty=2) + geom_vline(xintercept=2, lty=3, col="indianred", lwd=1) 

logistic_dens
```

--

- Same thing for a probit

```{r, echo=F, fig.width=5, fig.height = 3, fig.align="center"}
probit_dens <- ggplot() + xlim(-5, 5) + geom_function(fun=dnorm) + theme_bw() + ylab("Density") + geom_area(aes(x=c(-5,2)),stat = "function", fun = dnorm, fill = "dodgerblue", xlim = c(-5, 2)) + 
  xlab(expression(epsilon)) + geom_vline(xintercept=0, lty=2) + geom_vline(xintercept=2, lty=3, col="indianred", lwd=1) 

probit_dens
```

---

# Ordinal logit

- Using the latent variable formulation lets us extend the binary outcomes to logit to outcomes with multiple discrete levels.
- For example, consider an "ordinal" response variable
  - (e.g.) a likert scale (1 = "strongly disagree", 3 = "neutral", 5 = "strongly agree")
--

- The responses are *ordered* but the intervals between them do not have any meaning.
  - A popular modeling technique with an ordinal $Y$ is to treat it as an expression of a latent variable
--

- In an *ordered logit* model, an ordinal outcome $Y$ with $L$ unique, ordered values $Y \in \{1, 2, 3, \dotsc L\}$ is a function of a latent variable and a set of $L-1$ cutpoints $\kappa$

$$Y_i = \begin{cases}1 &\text{if } Y_i^* \le \kappa_1 \\
                      2 &\text{if } \kappa_1 < Y_i^* \le \kappa_2\\
                      3 &\text{if } \kappa_2 < Y_i^* \le \kappa_3\\
                      \vdots &\\
                      L &\text{if } Y_i^* > \kappa_{L-1}\end{cases}$$
--

- We still assume $Y_i^*$ has a logistic distribution

$$Y_i^* = X_i^{\prime}\beta + \epsilon_i$$

$$\epsilon_i \sim \text{Logistic}(0, 1)$$

---

# Ordinal logit

- To estimate the model, we now have to estimate both the $\beta$ parameters and the $L-1$ $\kappa$ parameters
  - Note we omit the intercept from our $X_i$ -- the $\kappa$ parameters are essentially the "intercepts" for each boundary between the choices
--

- How does the model map changes in the latent variable to changes in probability?

$$Pr(Y_i = 1) = Pr(Y_i^* \le \kappa_1) = Pr(\epsilon_i \le \kappa_1 - X_i^{\prime}\beta)$$
$$Pr(Y_i = 2) = Pr(Y_i = 2 \cup Y_i = 1) - Pr(Y_i = 1) = Pr(Y_i^* \le \kappa_2) - Pr(Y_i^* \le \kappa_1)$$
$$Pr(Y_i = 3) = Pr(Y_i = 3 \cup Y_i = 2 \cup Y_i = 1) -  Pr(Y_i = 2 \cup Y_i = 1)  = Pr(Y_i^* \le \kappa_3) - Pr(Y_i^* \le \kappa_2)$$
--

---

# Ordinal logit

- Let's think in terms of the distribution of $Y_i^*$ and the cut-points for a case with $L=3$. For $X_i^{\prime}\beta = 0$:

```{r, echo=F, fig.width=7, fig.height = 3, fig.align="center"}
logistic_dens <- ggplot() + xlim(-8, 8) + geom_function(fun=dlogis) + theme_bw() + ylab("Density") + geom_area(aes(x=c(-8,-1)),stat = "function", fun = dlogis, fill = "dodgerblue", xlim = c(-8, -1)) +  geom_area(aes(x=c(-1,2)),stat = "function", fun = dlogis, fill = "darkorange", xlim = c(-1, 2)) + geom_area(aes(x=c(2,8)),stat = "function", fun = dlogis, fill = "yellow", xlim = c(2, 8)) + 
  xlab("Y*") + geom_vline(xintercept=0, lty=2) + geom_vline(xintercept=-1, lty=3, col="darkgreen", lwd=1) + geom_vline(xintercept=2, lty=3, col="darkgreen", lwd=1) 

logistic_dens
```
--

- Shifting the mean of $Y^*$ up by 3:

```{r, echo=F, fig.width=7, fig.height = 3, fig.align="center"}
logistic_dens <- ggplot() + xlim(-8, 8) + geom_function(fun=dlogis, args=list(location=3)) + theme_bw() + ylab("Density") + geom_area(aes(x=c(-8,-1)),stat = "function", fun = dlogis, fill = "dodgerblue", xlim = c(-8, -1),  args=list(location=3)) +  geom_area(aes(x=c(-1,2)),stat = "function", fun = dlogis, fill = "darkorange", xlim = c(-1, 2),  args=list(location=3)) + geom_area(aes(x=c(2,8)),stat = "function", fun = dlogis, fill = "yellow", xlim = c(2, 8),  args=list(location=3)) + 
  xlab("Y*") + geom_vline(xintercept=3, lty=2) + geom_vline(xintercept=-1, lty=3, col="darkgreen", lwd=1) + geom_vline(xintercept=2, lty=3, col="darkgreen", lwd=1) 

logistic_dens
```

---

# Ordinal logit

- This structural mapping from $Y^*$ to $P(Y_i = l)$ is not trivial - it encodes a particular assumption about how these probabilities relate to one another.
- Referred to as the **proportional odds** assumption

$$\log\left(\frac{Pr(Y_i \le l)}{1 - Pr(Y_i \le l)}\right) = \kappa_l - X_i^{\prime}\beta$$
--

- **Intuitively** -- The $\beta$ parameters are shared across all the possible outcome levels. Information about whether someone responds "strongly agree" *also* tells us something about how likely they would be to respond specifically to "agree", "neutral", "disagree", and "strongly disagree"
  - Not trivial if there's non-monotonicity or heterogeneous effects. 
  - Be careful with these models!
  
---

class: title-slide

# "Robust" inference and misspecification

---

# Misspecification

- Often we'll encounter settings where we only believe *part* of the maximum likelihood model
--

- Perhaps that we've correctly specified $g(E[Y_i | X_i])$, but we don't believe the complete distributional assumptions we've placed on the outcome
  - We don't believe the observations are "**identically** distributed"
--

- Or we believe that some outcomes are correlated with one another even after conditioning on the covariates 
  - We don't believe they're **independently** distributed.
--

- What can we do?
  - Suppose our MLE is still **consistent** for something we're interested in (e.g. the CEF).
  - But the outcome model is incorrect (e.g. due to heteroskedasticity) and so our variance estimator is mis-specified.
  - Can we still estimate the variance of our MLE when it's *not* efficient? Yes!

---

# "Robust" Standard Errors

- Suppose we obtain a maximum likelihood estimate based on maximizing the log-likelihood

$$\ell(\theta|\mathbf{y}) = \sum_{i=1}^n \log f_i(y_i | \theta)$$

- Note that the density here is being indexed by $i$
  - We're relaxing the "identically" distributed assumption -- we're leaving the full distribution of $Y_i | \theta$ unspecified
--

- Denote the score

$$\ell^{\prime}(\theta|\mathbf{y}) = \sum_{i=1}^n \frac{\partial}{\partial \theta} \log f_i(Y_i | \theta) = \sum_{i=1}^n g_i(Y_i|\theta)$$
--

- And the Hessian

$$\ell^{\prime\prime}(\theta|\mathbf{y}) = \sum_{i=1}^n \frac{\partial^{2}}{\partial \theta^{2}} \log f_i(Y_i | \theta) = \sum_{i=1}^n h_i(Y_i|\theta)$$

---

# "Robust" Standard Errors

- Assume that there is a true $\theta_0$
- Let's do a Taylor approximation of the likelihood around that true value

$$\ell(\theta|\mathbf{y}) =  \ell(\theta_0|\mathbf{y}) + \ell^{\prime}(\theta_0|\mathbf{y})(\theta - \theta_0) + \frac{1}{2}(\theta- \theta_0)^{\prime}\ell^{\prime\prime}(\theta_0|\mathbf{y})(\theta-\theta_0)$$
--

- The MLE $\hat{\theta}$ is the solution that maximizes that likelihood function. 
  - So we know it satisfies the equation that sets the score equal to $0$. Let's take the derivative of the above w.r.t $\theta$ and solve for $0$
  
$$0 = \ell^{\prime}(\theta_0|\mathbf{y}) + (\hat{\theta} - \theta_0)^{\prime}\ell^{\prime\prime}(\theta_0|\mathbf{y})$$
--

- Solve for $\hat{\theta}$

$$(\hat{\theta} - \theta_0) = [-\ell^{\prime\prime}(\theta_0|\mathbf{y})]^{-1}\ell^{\prime}(\theta_0|\mathbf{y})$$
---

# "Robust" Standard Errors

- Plug in our definitions from before

$$(\hat{\theta} - \theta) = \bigg(- \sum_{i=1}^n h_i(Y_i|\theta_0)\bigg)^{-1}\bigg(\sum_{i=1}^n g_i(Y_i|\theta_0)\bigg)$$

--

- Multiplying by 1 and using properties of inverses

$$(\hat{\theta} - \theta) = \bigg(-\frac{1}{n}\sum_{i=1}^n h_i(Y_i|\theta_0)\bigg)^{-1}\bigg(\frac{1}{n}\sum_{i=1}^n g_i(Y_i|\theta_0)\bigg)$$

--

- We make two asymptotic arguments
  - The first term converges in probability to the Fisher information
  - The second term times $\sqrt{n}$ converges in distribution to a normal distribution with mean $0$ and variance $Var(g_i(Y_i|\theta_0))$

---

# "Robust" Standard Errors

- So asymptotically, we have

$$\sqrt{n}(\hat{\theta} - \theta) \underset{d}{\rightarrow}  \bigg[\mathbf{E}[-\ell^{\prime\prime}(\theta_0|\mathbf{y})]\bigg]^{-1}\mathcal{N}\bigg(0, Var(g_i(Y_i|\theta_0))\bigg)$$

--

- And therefore, the asymptotic variance of $\hat{\theta}$ is

$$Var(\hat{\theta}) = \bigg[\mathbf{E}[-\ell^{\prime\prime}(\theta_0|\mathbf{y})]\bigg]^{-1}\bigg[Var(g_i(Y_i|\theta_0))\bigg]\bigg[\mathbf{E}[-\ell^{\prime\prime}(\theta_0|\mathbf{y})]\bigg]^{-1}$$

---

# "Robust" Standard Errors

- We'll use our usual plug-in estimator for the Hessian at the MLE
  - But earlier, we showed that under the model, the middle term *also* equals the expected negative hessian (and therefore cancels with one of the inverses).
  - Instead, we'll use a different plug-in estimator using the outer product of the gradient of each observation at the MLE.

$$\widehat{Var(\hat{\theta})} = \bigg[-\sum_{i=1}^n h_i(Y_i|\hat{\theta})\bigg]^{-1}\bigg[\sum_{i=1}^n g_i(Y_i|\hat{\theta})  g_i(Y_i|\hat{\theta}) ^{\prime}\bigg]\bigg[-\sum_{i=1}^n h_i(Y_i|\hat{\theta})\bigg]^{-1}$$

---

# The "Sandwich" Estimator

- We call this the "sandwich" estimator because it consists of a "meat" between two slices of identical "bread"

$$\widehat{Var(\hat{\theta})} = \underbrace{\bigg[-\sum_{i=1}^n h_i(Y_i|\hat{\theta})\bigg]^{-1}}_{\text{bread}}\underbrace{\bigg[\sum_{i=1}^n g_i(Y_i|\hat{\theta})  g_i(Y_i|\hat{\theta}) ^{\prime}\bigg]}_{\text{meat}}\underbrace{\bigg[-\sum_{i=1}^n h_i(Y_i|\hat{\theta})\bigg]^{-1}}_{\text{bread}}$$

--

- These estimators are **consistent** for the asymptotic variance under mis-specification of the outcome distribution
  - But finite-sample properties can be not great -- typically we get *undercoverage* in small samples.
  - Typically add a finite-sample correction (e.g. $\frac{N}{N-K}$) 
  
---

# The regression "sandwich" 

- A familiar form of the sandwich arises from linear regression.
- Remember our OLS estimator

$$\hat{\beta} = (\mathbf{X}^{\prime}\mathbf{X})^{-1}(\mathbf{X}^{\prime}\mathbf{y})$$
--

- We wrote this as 

$$Var(\hat{\beta}) = Var\bigg((\mathbf{X}^{\prime}\mathbf{X})^{-1}(\mathbf{X}^{\prime}\epsilon)\bigg)$$
--

- And took the variance

$$\hat{\beta} = \beta + (\mathbf{X}^{\prime}\mathbf{X})^{-1}(\mathbf{X}^{\prime}\epsilon)$$

---

# The regression "sandwich"

- We're doing inference conditional on $\mathbf{X}$

$$Var(\hat{\beta}) = Var\bigg((\mathbf{X}^{\prime}\mathbf{X})^{-1}(\mathbf{X}^{\prime}\epsilon)\bigg)$$

$$Var(\hat{\beta}) = (\mathbf{X}^{\prime}\mathbf{X})^{-1}(\mathbf{X}^{\prime}Var(\epsilon)\mathbf{X})(\mathbf{X}^{\prime}\mathbf{X})^{-1}$$
--

- $Var(\epsilon)$ is the variance-covariance matrix of the errors.
  - Since the errors are mean zero, it's the expectation of the outer-product of the errors $Var(\epsilon) = E[\epsilon\epsilon^{\prime}]$
  - Under our Gauss-Markov assumptions (specifically, independence and homoskedasticity), $Var(\epsilon) = \sigma^2\mathbf{I}$. And so

---

# The regression "sandwich"

- Instead of making that assumption, we'll just plug in the observed residuals $\hat{\epsilon}$ as estimators of $\epsilon$
  - This yields the classic "robust" variance estimator:
  
$$\widehat{Var(\hat{\beta})} = \underbrace{(\mathbf{X}^{\prime}\mathbf{X})^{-1}}_{\text{bread}}\underbrace{(\mathbf{X}^{\prime}\hat{\epsilon}\hat{\epsilon}^{\prime}\mathbf{X})}_{\text{meat}}\underbrace{(\mathbf{X}^{\prime}\mathbf{X})^{-1}}_{\text{bread}}$$
---

# Implementing sandwich SEs

- Let's go back to our propensity score logistic regression of abciximab treatment on the predictive covariates

```{r}
library(sandwich) # this contains "estfun()"

logit_sandwich <- solve(-hessian(logit_mle))%*%(t(estfun(logit_mle))%*%estfun(logit_mle))%*%solve(-hessian(logit_mle))
```

- Compare the robust SEs to the ones we had earlier

```{r}
# Non-robust
logit_SEs
# Robust
logit_sandwich_SEs <- sqrt(diag(logit_sandwich))
logit_sandwich_SEs
```

---

# Implementing sandwich SEs

- These are also implemented in the `sandwich` package which operates on a lot of common MLE packages in R

```{r}
logit_sandwich_vcov <- sandwich::sandwich(logit_mle)
sqrt(diag(logit_sandwich_vcov))
logit_sandwich_SEs
```

---

# The Bootstrap

- An alternative approach to estimating the sampling variance of any estimator via monte carlo methods is the "bootstrap"
- **Intuitively**: Bootstrap methods use the *empirical* distribution of some random variable to estimate its unknown *theoretical* distribution
  - Use **monte carlo** methods to approximate that empirical distribution.
--

- For regression, it is common to use the **pairs bootstrap** to obtain a heteroskedasticity-robust variance estimator (Freedman, 1981)
  1. Start with the original sample $(Y_1, X_1), (Y_2, X_2), \dotsc, (Y_n, X_n)$
  2. Generate a new "bootstrapped" sample of size $n$: $(Y_1^*, X_1^*), (Y_2^*, X_2^*), \dotsc, (Y_n^*, X_n^*)$ where the probability $Pr[(Y_j^*, X_j^*) = (Y_i, X_i)] = \frac{1}{n}$ for each $i$ and $j$ (sampling *with* replacement)
  3. Compute $\hat{\beta}^*$ using the boostrapped sample $(Y_1^*, X_1^*), (Y_2^*, X_2^*), \dotsc, (Y_n^*, X_n^*)$
  4. Repeat 2-3 a large number of times $B$ to obtain a sample of bootstrapped estimates $\hat{\beta}^{(1)}, \hat{\beta}^{(2)}, \dotsc, \hat{\beta}^{(B)}$
  5. Use the sample moments of the bootstrapped estimates to estimate the quantities of interest (e.g. standard error; percentile intervals)

---

# Implementing the Bootstrap

- Let's do a pairs/resampling bootstrap 

```{r, cache=T}
set.seed(60637)
nIter <- 1000
beta_boot <- matrix(nrow=nIter, ncol=ncol(X_mat))
for (i in 1:nIter){
  boot_rows <- sample(1:nrow(X_mat), replace=T)#Resample rows 
  mle_boot <- maxLik(logit_loglik, 
                      Y=pci$abcix[boot_rows],
                      X=X_mat[boot_rows,],
                      start = rep(0, ncol(X_mat)),
                      method = "NR")
  beta_boot[i,] <- coef(mle_boot)
}
```

---

# Implementing the Bootstrap

- Compare our bootstrap variance-covariance matrix to the sandwich estimator

```{r}
# Sandwich
logit_sandwich_SEs
# Bootstrap
boot_SEs <- sqrt(diag(var(beta_boot)))
boot_SEs
```

---